---
title: "HW1-33A"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Call Expressions

### The Normal Curve

The standard normal curve is centered at 0 and has a spread, AKA standard deviation, of 1. This curve is drawn in the figure below along with a normal curve that has a center of 2 and a spread of 1.5 (dashed curve).

```{r}
curve(dnorm, from = -3, to = 6, ylab = "Density")
curve(dnorm(x, mean = 2, sd = 1.5), add = TRUE, lty = 2)
```


The family of normal curves all have the same shape and each is uniquely specified by its center (mean) and spread (standard deviation)  


1. Read the documentation for `rnorm()` to figure out how to call `rnorm()` to generate 5 random values from a normal distribution with mean 17 and standard deviation 3.
```{r}
rnorm(5, mean = 17, sd = 3)
```

2. Create a vector called `normSamps` containing 1000 random values from a normal distribution with mean 1 and standard deviation 2.
```{r}
# Replace the 1 on the right hand side of the equation with your code. Do not modify the call to set.seed()
set.seed(66)
normSamps = rnorm(1000, mean = 1, sd = 2)
```

3. The 10% trimmed mean is obtained by taking the mean of the middle 80% of the values (removing the lowest 10% and the highest 10% of the values). Find the 10% trimmed mean of normSamps and assign this value to `mean10`. 
Hint: Look at the help file for the mean() function.
```{r}
# Replace the 1 on the right hand side of the equation with your code.
mean10 = mean(normSamps, trim = 0.10)
mean10
```

4. Use `sum()` to calculate the fraction of values in 
`normSamps` that are less than 3. 
Write your solution so that it coerces a logical vector to numeric without using the function as.numeric().
Write your code generically so that it doesn't depend on
`normSamps` being a vector of 1000 elements.
Assign the fraction to the variable `frac3`
Hint: consider using the `length()` function.
```{r}
# Replace the 1 on the right hand side of the equation with your code
frac3 = sum(normSamps < 3)/length(normSamps)
frac3
```



5. A percentile is the value at which a given percentage of samples fall at or below it and is sometimes synonomously called a quantile, e.g., the 90th percentile is the value at which 90% of data values fall at or below it. Calculate the 90th percentile of `normSamps` and assign this value to `per90`. 
Hint: The `quantile()` function is helpful for this question.
```{r}
# Replace the 1 on the right hand side of the equation with your code
per90 = quantile(normSamps, probs = 0.90)
per90
```



6. Calculate the greatest absolute distance from 1 for the values in `normSamps`. Assign this value to a variable called `maxDev`, short for "maximum deviation" from the mean. Hint: The functions abs() and max() are helpful for this question.
```{r}
# Replace the 1 on the right hand side of the equation with your code
maxDev = max(abs(normSamps - 1))
maxDev
```
