---
title: 'US Presidential Elections: A Case Study'
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read the file 

The file is located on the web at
http://www.stat.berkeley.edu/users/nolan/data/usPres.txt

Examine the contents of the file and determine the format.

+ Is it plain text?
+ Is it delimited? What are the delimiters?
+ Does it have a header?

Use the answers to these questions to write code to
read the file into *R*.
Assign the return value to `usp`.

```{r readit}
usp =  read.csv(url("http://www.stat.berkeley.edu/users/nolan/data/usPres.txt"),
                header = TRUE)
```

## Check the data frame

Examine the data frame.

+ What are the number of rows and columns?
+ What are the names of the variables?
+ Do the first few rows look ok? What about the last few?

```{r checkit}
dim(usp)
head(usp)
tail(usp)
```


Look at summary statistics for each of the variables.
Does everything look in order?

```{r}
summary(usp)  #run this code
```

## Check Data Types

Confirm that the variables that are factors should be.

```{r changeType}
sapply(usp, class)  #run this code
```

Should the names of the winners and losers be factor or character?

Convert them to character strings and reassign to the data frame.
Replace the existing variables with the new ones. That is,
the end result should have the same number of vectors.
```{r}
usp$Win = as.character(usp$Win)
usp$Lose = as.character(usp$Lose)
sapply(usp, class)
```


## Explore

Below are five questions for exploration.
Use the `filter()`, `select()` and `slice()` functions in the 
`dplyer` package to help you.
Also consider the `table()`, `mean()` and other simple summary functions.
Assign the answer for each question to Expn, for n = 1, 2, 3, 4, 5.

```{r}
require(dplyr)
```


1. Names of Presidents who did not get over 50% of the vote.
```{r}
Exp1 = select( filter(usp, percent < 0.5), Win)

Exp1
```


2. Year and name of presidents who lost popular vote, but won the election
```{r}
Exp2 = select( filter(usp, margin < 0), Yr, Win)

Exp2
```


3. Party and names of candiate who won the popular vote, but lost the election
```{r}
Exp3 = select( filter(usp, margin < 0), Yr, Lose)

Exp3
```


4. Proportion of times Dems vs Reps won the election
```{r}
Exp4 = table(usp$win_party)/nrow(usp)

Exp4
```

5. Presidents who only served one term

```{r}

countWins = table(usp$Win)
countWins = as.data.frame(countWins)

Exp5 =filter(countWins, Freq == 1)

Exp5
```

