---
title: 'Reshaping Data Frames'
author: 
output:
  pdf_document: default
  html_document: default
  word_document: default
---

Load the following libraries - 
```{r, eval=TRUE, message=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
```

## Introduction to the Data

In this homework you will work with the data collected by loop detectors at a particular location of eastbound Interstate 80 in Sacramento, California, from March 14-20, 2003. There are three lanes - lane 1 is the leftmost, lane 2 is the center, and lane 3 is the rightmost lane. For each freeway lane, the flow and the occupancy were recorded in successive five minute intervals on the loop throughout the day. 
  
Flow is number of cars that went over the loop in a five minute interval and occupancy is the percentage of time there was a car over the loop in that same interval. The occupancy variable can be viewed as a measure of congestion.
 
 
Load the RDA file in to your R session.
```{r}
load(url("http://www.stat.berkeley.edu/users/nolan/data/traffic33A.rda"))
```

Let's take a look at the data 
```{r}
head(traffic)
```
How many variables are there? Can you tell what each variable stands for? Examine the summary statistics for each variable too.

```{r}
summary(traffic)
```

## Working with dates

To help us work with the dates, install the `lubridate` package so that you can load it into your R session.
```{r error=TRUE, message=FALSE}
library(lubridate)
```

The `lubridate` package provides functions to get components of date-time, such as `year()`, `month()`, etc. With the help of the functions in this package, create two more columns in the traffic dataframe 
   - 'day' containing the day part of 'time'
   - 'hour' indicating the hour of the day
Both these new columns shoud be factors (use the `factor()` function).
```{r}
traffic$day = 1
traffic$hour = 1
```
  
Check the dimensions of the data frame to confirm that there are two more columns
```{r}
dim(traffic)
```
 
The `day` column should range from 14 to 20 as the traffic dataframe contains measurements from 14th to 20th March. 
The hour should have values from 0 to 23.
```{r}
table(traffic$day)
table(traffic$hour)
```

Do you notice something unexpected from the frequency of the 'day' column? The number of 5 minute intervals in a day is 288 (24*12),
but on 20th there seems to be only 12 entries. 
The data on 20th March is not complete. Remove this day from the dataframe. Make sure to use quotations in your logical expression while filtering.  Assign this new data frame to `traffic2`.
```{r}
traffic2 = 1
```

## Gathering the data

Gather the occupancies for the three lanes of traffic in `traffic2` together into one variable (`occ1`, `occ2` and `occ3`). The new data frame should have the following features:

+ The flow measurements (`flow1`, `flow2` and `flow3`) should be removed.
+ The `day` and `hour` variables should be kept.
+ The occupancy values should be collected into the variable `occ`.
+ The key should be called "lane"
+ The new data frame should be called `trafficLong`
```{r}
trafficLong = 1
```

Now gather just the flow measurements (`flow1`, `flow2` and `flow3`).
This data frame should:

+ have a column called `flow`
+ have a column called `lane`
+ be called `flowLong`

```{r}
flowLong = 1
```

Add `flowLong` to `trafficLong`.
```{r, error = TRUE}
trafficLong$flow  = flowLong$flow
```



Convert the `lane` column (that you created in the above steps) into a factor and rename the levels as left, middle and right.  Use the `factor()` and `level()` functions to do this.
```{r}

```

## Plots 

Now that we have gathered the data, we will use this data to make some graphs. Make sure you use the `trafficLong` data. You will notice how gathering the data into a long format makes it easier to plot. 

### Plot #1 
Augment the density curve below so that:

+ there are curves for each lange
+ log-scale is used on the x-axis
+ label the x-axis and add a title to the plot
+ use the black and white theme.

```{r, error = TRUE}
ggplot(data = trafficLong) +
  geom_density(aes(x = flow))
```



### Plot #2 
Augment the scatterplot below to:

+ use color to distinguish the lanes
+ add smooth curves for each lane
+ label the x and y axis

```{r, error = TRUE}
ggplot(data = trafficLong) +
  geom_point(aes(x = occ, y = flow)) +
  geom_smooth(aes(), 
              method = "loess", span = 0.3, se = FALSE)
```



### Plot #3 
Plot boxplots of flow for every hour and lane combination.
Add x and y axis labels.
```{r, error = TRUE}
ggplot(data = trafficLong)

```


