---
title: 'Homework 7: Case Study - Food Safety, Part 2'
author: "A solution"
output: pdf_document
---

# Introduction

We continue with the analysis of the restaurant inspection scores from
HW 6. In this HW we investigate the specific violations and try to 
associate different sorts of violations with scores.

Rather than repeating all of our earlier work of cleaning, grouping,
and joining the business and inspection data, we begin where we left off
in HW 6 and load the data frame created.
The call to `load()` below loads in a data frame that should be the same as (or close to) the one you created in HW 6. In addition, we read in the violations raw data file.
```{r}
load(url("http://www.stat.berkeley.edu/users/nolan/data/rest16.rda"))

vio = read.csv(url("http://www.stat.berkeley.edu/users/nolan/data/violations.csv"), 
               stringsAsFactors = FALSE)
```

We will also need the `checkGran()`  function that you wrote in HW 6.
Copy it into the code chunk below.
```{r}
checkGran = function(key, dups = FALSE) {
  
  num = nrow(key)
  dupLoc = duplicated(key)
  uniq = sum(!dupLoc)
  
  info = c(num = num, uniq = uniq)

  if (dups) {
    dupLoc = which(dupLoc)
    info = c(info, dupLoc)
  } 
  
  return(info)
}
```



Load the libraries: ggplot2, dplyr and tidyr.
```{r,echo=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(gridExtra)
```


# Violations Data

Confirm the names of the variables in `vio` with a call to  `names()`:

```{r}
names(vio)
```

Notice that we have only 3 variables in violations data frame. We are familiar with the business id and the date of the inspection. 

As with the first two data frames, we want to determine the granularity, i.e.,
we want to ascertain what uniquely defines the records.

Use your `checkGran()` function from HW 6, and call it three times:

+ with `business_id` as the key
+ with `business_id` and `date`
+ with `business_id` and `date` and `description`

```{r}
checkGran(vio["business_id"])
checkGran(vio[c("business_id", "date")])
checkGran(vio[c("business_id", "date", "description")])
```

### Q. Doess the combination business_id and inspection date uniquely identify the records?

ANS: No, there can be multiple violations to an inspection.


Examine the first 20 records of `vio`.

```{r}
head(vio, 10)
```

Just visually inspecting the first few records in `vio` we see that each inspection has multiple entries. Reading the descriptions, we see that if corrected, a date is listed in the description within square brackets. This date appears to be the same date as the inspection.

## Narrow down the data

Recall that our data frame `ins.bus2016` has information only about 
the inspections in 2016. 
Narrow the violations data to contain information only on the inspections in `ins.bus2016`. 

To do this, use the `semi_join()` function. 
Specifically, use `business id` and `date` from `ins.bus2016` to drop the violations from `vio` for other years and inspections. Call this new data frame `vio2016`.

```{r}
vio2016 = semi_join(vio, ins.bus2016, by = c("business_id", "date"))
```


## Number of violations

As we explore data, we often create new variables that we think might be useful in our analysis, such as the number of violations per inspection.

To see if this variable may be of interest, call `count()` and assign it to `num_vio`.
                   
```{r}
num_vio = count(vio2016, business_id)
head(num_vio)
```



Make a histogram of `n` in `num_vio`.

```{r}
ggplot(num_vio, aes(x = n)) +
  geom_histogram()
```

### Q. Describe the distribution of the number of violations.

+ The number of violations has a long/short left/right tail
+ Most inspections have xx to yy violations.
+ Inspections with more than xx violations are rare 
+ Inspections with xx violations are not shown in this histogram.


## Summary of findings

Let's take stock of what we have learned about the violations data and what might be the next steps in our investigation. We found that:

* Each record corresponds to one violation that was found on an inspection. 
* When a violation is corrected this information is provided in the description field.
* The distribution of the number of violations shows that some restaurants had many violations. 


# Examine the violations

In order to gain more insight into the violations, 
we will need to derive numeric variables from the violation description. 
For example, we can derive logical variables that indicate:

* Whether the violation was corrected.
* Whether the violoation is high risk
* Whether the violation involved the absence/presence of, say, vermin, human cleanliness, floor cleanliness, etc.


## Corrected violations

We noticed earlier that when a violation is corrected, the description includes information about when it was corrected, e.g. [ date violation corrected: 9/1/2016 ].  

We have not seen the square brackets used otherwise in a description so
we can write code to look for a square bracket in the description and use
it to indicate whether the violation was corrected.

We can use the function `grepl()` to search for a left square bracket `[` in the description of a violation.
Let's try it out. We set up a test vector or 3 descriptions with:
```{r}
testd = vio2016$description[1:3]
testd
```

Then we search in them for "[" with
```{r}
grepl("\\[", testd)
```

Create a new variable called `corrected` in `vio2016` that is a logical 
vector that is TRUE if the violation was corrected and FALSE if not.
```{r}
vio2016$corrected = grepl("\\[", vio2016$description)
```

### Q. How many violations were corrected? not corrected?

```{r}
sum(vio2016$corrected)
sum(!vio2016$corrected)
```

## Cleaning the description

Before we continue deriving variables, we clean the descriptions so that it will be easier to find patterns.  
The code below converts any upper case letters to lower case letters.
That way we don't have to look for "High risk" and "high risk".
It also eliminates extra blanks and the corrected information so that 
the descriptions are more uniform.

```{r}
vio2016$description = tolower(vio2016$description)
vio2016$description = sub("[[:blank:]]{2,}", "", vio2016$description)
vio2016$description = sub("\\[.*\\]", "", vio2016$description)
```


## Risk of a violation

Use `grepl()` to create a logical vector, called `hi_risk`, according to whether or not the term "high risk" appears in the description.
This vector should be added to `vio2016`.

```{r}
vio2016$hi_risk = grepl("high risk", vio2016$description)
```

### Q. How many high risk violations are there? 

```{r}
sum(vio2016$hi_risk)
```


## Common and rare violations

Now that the violations are more uniform in appearance, we can 
table them and examine the most and least common. 


Use `count()` to create a table of counts of the descriptions in `vio2016` and call it `vioTable`.

```{r}
vioTable = count(vio2016, description)
```

The following call will arrange the table in order according to the counts and display the top 25.
```{r}
head(arrange(vioTable, desc(n)),25)
```

### Q. How many unique violation descriptions are there?

```{r}
dim(vioTable)
```

ANS: 67

Also, examine the 20th to the 30th most common.
```{r}
arrange(vioTable, desc(n))[20:30, ]
```


And, examine the 10 least common.

```{r}
head(arrange(vioTable, n),10)
```


## Derive violation indicators

Create at least 8 indicator variables (logical vectors) that indicate a certain kind of violation. Try to have some of these variables group descriptions with a common theme together. For example, does the desctiption mention rats, mice, or other vermin? Is it related to the cleanliness of the workers in the restaurant? Or the facilities?


To begin create a list of terms, where the list is of length >=8, and each element in the list is a character vector of themed words. Call this list `vio_types` and name each element according to the type of violation that it represents. Use your investigations of `vioTable` to create your vectors of words. You may want to reexamine the descriptions in `vioTable` more carefully as you go.

We created a short list of two character vectors for you as a starting point. You may add more terms to those two vectors, and you are to create at least 6 more vectors.

```{r}
vio_types = list(
  unclean = c("unsanitary", "unclean", "clean"),
  human = c("hand", "hair", "nail"),
  surface = c("floor", "surface", "wall"),
  food = c("food", "fish", "oyster"),
  vermin = "vermin",
  facilities = c("facilities", "sewage", "plumbing"),
  methods = c("method", "thermometer", "heat", "utensil")
)
```

Next, write a function called `search_terms()` that will search 
in the descriptions for any of the terms in a character vector.

The inputs and output to this function are:

+ `x`: required vector of descriptions to search in
+ `terms`: required vector of words/terms to search for
+ logical vector the same length as `x` in which an element is TRUE if any of the terms were found in the corresponding element of `x`

You will need to use `grepl()` and a for loop. One approach is the following:

+ set up a logical vector of the correct length to start (use `vector()`)
+ loop over each term in `terms`
+ for each execution of the loop, update the logical vector according to which elements of `x` contain the term

```{r}
search_terms = function(x, terms){
  
  out = vector(mode = "logical", length = length(x))
  
  for (i in 1:length(terms)) {
     out = out | grepl(terms[i], x)
  }
  return(out)
}
```

Test your function with the following code.

```{r}
x = c("hi there", "bye there", "bye cat for now", "it is a cat")

search_terms(x, c("cat", "bye"))   #returns FALSE  TRUE  TRUE  TRUE
search_terms(x, c("z", "now", "cat")) #returns FALSE FALSE  TRUE  TRUE
search_terms(x, c("there"))    #returns TRUE  TRUE FALSE FALSE
search_terms(x, c("z", "abc"))  #returns FALSE FALSE FALSE FALSE 
```


Create your indicator variables with a call to `lapply()`
```{r}
dummies = lapply(vio_types, search_terms, x = vio2016$description)
dummies = as.data.frame(dummies)
```

Check that your variables have a reasonable number of TRUEs,
```{r}
sapply(dummies, sum)
```


Add your indicator variables to the `vio2016` data frame with a call to `bind_cols()`.

```{r}
vio2016_all = bind_cols(vio2016, dummies)
```


##  Collapsing violations

In order to join the violations data with the inspection data, we need
to aggregate the violations data up to the inspection level.

Use `group_by()` and `summarize()` to do this. The grouped and summarized data should have the following variables:

+ `business_id`
+ `date`
+ `num_vio`: the number of violations in the inspection
+ `hi_risk`: the number of high risk violations in the inspection
+ `corrected`: the number of corrected violations in the inspection
+ `unclean`: TRUE if any of the violations had one of the unclean terms
+ `human`: TRUE if any of the violations had one of the human terms
+ plus at least 6 additional TRUE/FALSE variables 

Call this aggregated data frame, `vio2016agg`
```{r}
vio2016agg = group_by(vio2016_all, business_id, date)
vio2016agg = summarize(vio2016agg, hi_risk = sum(hi_risk), 
                       corrected = sum(corrected),
                       unclean = any(unclean), human = any(human), 
                       surface = any(surface), food = any(food), 
                       vermin = any(vermin), facilities = any(facilities),
                       methods = any(methods), 
                       num_vio = n())
```


## Joining violations and inspections

Join `vio2016agg` with the inspection data frame `ins2016M.B`. 
You will want to keep all the records in the inspection data frame. 
Call this new data frame, `ins.vio2016`

```{r}
ins.vio2016 = left_join(ins.bus2016, vio2016agg, 
                         by = c("business_id", "date"))
```

Note that some restaurants were inspected and no violations listed,
even when the score is not perfect. These records will have NA in all of the
violation variables. Will want to replace some of the values in these variables, especially the `num_vio` should be set to 0. And the logicals should be set to FALSE.  The `replace_na()` function does this.
Complete the code below to replace all NAs in `num_vio`, `hi_risk`, `corrected`, `unclean`, `human`, and the rest of your indicator variables.

```{r}
ins.vio2016 = replace_na(ins.vio2016, 
                  list(num_vio =0, hi_risk = 0, corrected = 0,
                       unclean = FALSE, human = FALSE, surface = FALSE,
                       food = FALSE, vermin = FALSE, facilities = FALSE,
                       methods = FALSE))
```

# Exploring violations and scores

Now that we have combined the data, we can explore the relationship 
between the violations and inspection scores.

## Score and Number of violations

Make a scatterplot of the inspection score and number of violations 
+ the number of violations and the number of corrections for 
restaurants with multiple violations


```{r}
ggplot(ins.vio2016, aes(x = score, y = num_vio)) +
  geom_point()
```

## Corrections and number of violations

Make a scatterplot of the number of violations and the number of corrections for restaurants with multiple violations.

```{r}
ggplot(filter(ins.vio2016, num_vio > 0), 
       aes(x = num_vio, y = corrected)) +
  geom_point()
```

## Score and high risk violations

Make side-by-side boxplots of the score versus the number of high risk violations.

```{r}
ggplot(ins.vio2016, 
       aes(x = factor(hi_risk), y = score)) +
  geom_boxplot()
```


### Q. Roughly how do 1 and 2 high risk violations impact the score?

ANS: One high-risk drops the median score by about 10 points and a second
one drops the median another 15 points.

## Impact of a violation

Make boxplots for all of the violation types that you have created.
Use the code below as a starting point. Try to arrange the boxplots
in order of the median score for the TRUEs.

```{r}
p0 = ggplot(ins.vio2016, aes(y = score))
p1 = p0 + geom_boxplot(aes(x = unclean))
p2 = p0 + geom_boxplot(aes(x = vermin))
p3 = p0 + geom_boxplot(aes(x = human))
p4 = p0 + geom_boxplot(aes(x = food))
p5 = p0 + geom_boxplot(aes(x = surface))
p6 = p0 + geom_boxplot(aes(x = methods))
p7 = p0 + geom_boxplot(aes(x = facilities))

grid.arrange(p4, p1, p2, p3, p6, p5, p7, ncol=4)
```

### Q. Which three violations seem to impact the score the most? 

ANS: Vermin, Human, and Food

## Combinations of violation types

Below is a set of boxplots that examine the impact on the score of 3 different 
types of violations. 

```{r}
ggplot(ins.vio2016, aes(y = score)) + 
  geom_boxplot(aes(x = vermin, fill = food)) +
  facet_wrap(~ human)
```

Make another combination of boxplots for 3 types of violations.

```{r}


```

### Q. Were you able to uncover a combination of violation impacts the score a lot?

ANS: 


