---
title: 'Homework 6: Case Study - Food Safety, Part 1'
author: "A Solution"
date: "Nov 8 2019"
output:
  pdf_document: default
---

# Introduction

In HWs 6 & 7, you will investigate restaurant food safety scores for restaurants in San Francisco. 
The scores and violation information have been made available by the San Francisco Department of Public Health. 
We will walk through the steps of data cleaning, merging, and exploratoration.
As we clean and explore these data, you will:

* Read simple csv files
* Work with data at different levels of granularity
* Clean data, e.g., remove records, modify variable values
* Group and join data frames
* Create new variables, e.g., indicators for certain types of food safety violations
* Explore characteristics and distributions of variables

Begin by loading the `ggplot2`, `dplyr` and `tidyr` libraries.

```{r,echo=FALSE,collapse=TRUE}
library(tidyr)
library(dplyr)
library(ggplot2)
```

Through out HWs 6 and 7, in addition to writing code in code blocks, we will ask you questions, which are to be answered in text. We identify the places where we are asking you to write a response to a question or questions with a capital Q. Be sure to answer all of these questions.


# Reading Food Safety Data

There are 3 files:

* http://www.stat.berkeley.edu/users/nolan/data/businesses.csv
* http://www.stat.berkeley.edu/users/nolan/data/inspections.csv
* http://www.stat.berkeley.edu/users/nolan/data/violations.csv
 
Read each into R. Use `read.csv()` to do this. Assign the businesses file to 'bus', the inpections file to 'ins' and the violations to 'vio'. Note that the business file has an 'ISO-8859-1' encoding so you will want to add the argument `encoding='ISO-8859-1'` to the call to `read.csv()`.
,
```{r,error=TRUE}
bus = read.csv( , stringsAsFactors = FALSE)
ins = read.csv( , stringsAsFactors = FALSE)
vio = read.csv( , stringsAsFactors = FALSE)
```

Examine a portion of each of the three data frames using `head()`.

```{r}


```

Find the dimension of each data frame:
```{r}


```

### Q. Why do the data frames have different numbers of rows?

You may want to refer to the class lecture materials from lecture #11 to help answer this question.

ANS: 


We will explore each data frame in turn, including determining its granularity and primary keys and exploring some of the variables indivdually. Let's begin with the businesses file, which has been read into the data frame `bus`.

## Prepare to investigate granularity

To confirm the granularity of a data frame we would like to compare the 
number of records with the number of unique values for a variable or a combination of variables. The *key* is a variable or combination of variables that uniquely identify each record in the data frame. 

### Q. From the variable names alone what do you think the key is for the bus and ins data frames?  

Not that a key can be a combination of variables.

* The key for `bus` is most likely:
* The key for `ins` is most likely:

To help confirm what the key is, write a function called `checkGran()` 
with the following inputs and output:

+ `key`: a required data frame that contains the potential key variables (and no other variables)
+ `dups`: a logical (with a default value FALSE) that indicates whether to return the duplicate rows or not. 
+ output - a named vector with names
    + `num` - the number of rows in the `key` data frame
    + `uniq` - the number of unique records in the data frame
    + when `dups` is TRUE, also return the indices of any duplicate records (these elements are not named)

Use the `norw()` and `duplicated()` and `which()` function in `checkGran()`

```{r}
checkGran = function(key, dups = FALSE) {
  

}
```

Test your function with the following data frame
```{r}
xx = data.frame(x = c(1, 1, 3, 4, 4), y = c(2, 2, 4, 5, 5), z = 1:5)
xx
```

The return values from the following calls to `checkGran()` are provided:
```{r}
checkGran(xx) # 5 5 named num and uniq

checkGran(xx, TRUE) # 5 5 named num and uniq

checkGran(xx[c("x", "y")])  # 5 3 named num and uniq

checkGran(xx[c("x", "y")], TRUE) #5 3 2 5 named num and uniq and two unnamed
```



# Examine the Business data 

From its name alone, we expect the `businesses.csv` file to contain information about the restaurants. Let's investigate the granularity of this dataset.

Is there one record per business ID?
Use the function that you wrote in the previous section to 
confirm that the number of rows matches the number of unique business IDs.

```{r}
checkGran( )  # 6315 6315
```


Check the uniqueness of the restaurant names. 
Call `checkGrabn()` to see if the name of the restaurant could also be a key.

```{r}
checkGran( )  # 6315 5692
```

### Q. Could the restaurant name be used as the key? Why or why not?

ANS: 

### Q. Even if the restaraunt names are unique, why is it not a good idea to use it as a key?

ANS: 

### Q. What variable(s) determines the granularity of the businesses data frame,

That is, what is the unique identifier for records in `bus`?

ANS:



## Zip codes

Next, let's  explore the zip code variable in the business data frame (it's called `postal_code` in the data frame). 


### Q. What data type is the zip code?


```{r}

```

ANS:


Use `unique()` to determine how many different postal codes there are in the data frame, and their values.

```{r}


```


### Q. Do you see any problems with the values of this variable?


ANS: 


Run the following code to change extended codes (more than 5 digits) to
5-digit zipcodes. 
```{r,error=TRUE}
bus$zip = substr(bus$postal_code, 1, 5)
```


Clean up the zip codes in `zip` by replacing bad values with NAs. 
You may find `%in%` useful here.

```{r}


```




## One restaurant zip codes

Often times, we need to get additional information in order to understand whether or not the data are bad values or simply unusual values. With zip codes, we can use the Internet to investigate the zip codes that have only a few businesses. 

### Q. Go online to investigate the zip codes: *94545*, *94602*, *94609*.  

Where are these zip codes?

+ 94545 is in:
+ 94602 is in:
+ 94609 is in:


Print out the records in `bus` for these zip codes. The `filter()` and `%in%` functions might be helpful. 
```{r}


```

### Q. Explain whether these strange zip codes might be appropriate or are typos. 

ANS:




Modify `zip' to replace 94602 with the 94102.
```{r}


```


Let's reduce our analysis to those zip codes that are in SF and that have many businesses. Below is a list of these zip codes. 

```{r}
validZip = c("94102", "94103", "94104", "94105", "94107",
"94108", "94109", "94110", "94111", "94112", "94114","94115",
"94116", "94117", "94118", "94121", "94122", "94123", "94124",
"94127", "94131", "94132", "94133", "94134")
```

Use the `%in%` function to filter `bus` to create a data frame with only these "valid" zip codes. Call this new data frame `busR`.

```{r}
busR = 1
```


## Summary of the business data

Before we move on to explore the other data frames, let's take stock of what we have learned and the implications of our findings on future analysis. 

* We found that the business id is unique across records so we can use it as a key to join tables. 
* We found that there are some bad values in zip code and zip codes outside the city. As a result, we focussed only on zip codes with many restaurants. 
(At another time, we may want to take the time to look up the restaurant address online and fix these errors.)   

Note that if we were interested in making maps and exploring geographic patterns, we would want to examine and clean the latitude and longitude values. 



# Investigate the inspection data

Let's now turn to the inspection data frame. Earlier, we found that `ins` has 4 columns, these are named `business_id`, `score`, `date` and `type`.  In this section, we determine the granularity of `ins` and investigate the kinds of information provided for the inspections. 

## Granularity

Use `checkGran()` to determine the granularity of `ins`.
Check whether or not it is `business_id` or the combined variables:
`business_id` and `date`.
```{r}
checkGran( )
```


You may find that the number of rows in `ins` and the unique values for the
key don't quite match. They are off by 2. Use the `dups` argument to `checkGran()` to determine which two records are duplicates of other records in the file.

Print out 5 records before and 5 after the first duplicate record (use `slice()` to do this.)
```{r}


```

### Q. How many times was this restaurant inspected? On which dates? What was the score on the duplicate date?

ANS: 


## Multiple inspections

Use `count()` to tally the number of inspections per restaurant.
Assign the return value from `count()` to `num_ins`.
```{r}
num_ins = 1

head(num_ins)
```

Make a histogram of the number of inspections that each business had.
For this plot make sure that:

+ y-scale is density, counts 
+ label the x-axis
+ bins are 1 wide
+ the fill and border color are grey and black, respectively

```{r}


```

### Q. Describe the distribution of the number of inspections.

To do this, modify the following statements to reflect the histogram.
That is, choose between the options separated by a slash and replace xx, yy, and zz with numeric values.

+ The scores have a long/short right tail
+ Most restaurants were inspected xx times in this period.
+ The greatest number of inspections was xx.

## Date of inspection

We would like to have the inspection date in format that makes analysis easy.
For example, we saw in the previous section that the inspection dates ranged
from 2014 to 2016. In this section, you will create a variable to contain
the year of inspection and a variable to contain the month of inspection.
We will use the year to reduce the file to include only inspections from 2016.

To help you, consider the following calls to the function `as.Date()`.

```{r,error=TRUE}
newDate = as.Date(as.character(ins$date), "%Y%m%d")

head(format(newDate, "%Y"))
head(format(newDate, "%b"))
```

Write a function called `partDate` with the following inputs and outputs:

+ `dates` a required vector of dates in the same format as `date` in `ins`
+ `part` a string that is either "year" or "month" that indicates whether the month or year is to be extracted from `dates`. The default is "year"
+ the return value is a vector the same length as `dates` that contains the year or the month 


```{r}
partDate = function( ) {
  
}
```

Call your function to add two new variables to `ins`, the variable `yr` and
the variable `mon` that hold the year and month, respectively.

```{r}


```


### Q. What range of years is covered in `ins`? Are there roughly same number of inspections each year?

Use `count()` to answer these questions.

```{r}

```

ANS:




Make a bar plot of the number of inspections in each month. 

+ Label the x axis. 
+ Make sure the months are in order, e.g. Jan is first followed by Feb. (one way to do this is to convert `mon` to a factor and specify `ordered=TRUE` and provide the levels argument as "Jan", "Feb", etc.)
+ Use only the data for 2014.


```{r}


```

### Q. Describe the distribution of inspection dates.

+ There are/are not roughly the same number of inspections per month 
+ The fewest number of inspections were in Jan/Feb/.../Nov/Dec.
+ More inspections in the spring/summer/fall/winter.



## Reduce the inspections to 2016

Let's examine only the inspections for one year, 2016. 
This puts businesses on a more equal footing.
Call this reduced data frame `ins2016`.

```{r}
ins2016 = 1
```


## Explore inspection score

What does the distribution of inspection scores in 2016 look like? 
From inspecting a few records, the inspection scores appear to be integer values. Make a histogram of the scores with a bin width of 1. 


```{r}


```

### Q. Describe the distribution of scores.

+ The scores have a long/short left tail
+ Scores over 90 are typically even/odd values
+ There are no scores of xx, yy, or zz.
+ Most scores are above xx.



## Assessing Granularity for 2016 inspections

Use `checkGran()` to confirm the granularity of `ins2016`.
Call the function with `dups = TRUE`.
```{r}
checkGran( )
```

### Q. Is there one record per key? 

ANS: 

Explore the distribution of the number of inspections that a restaurant has in a year? To answer this question, nest two calls to `count()`.

```{r, error=TRUE}
count(count(ins2016, xxx), n)
```

### Q. How many restaurants had 3 inspections in 2016? Did any have more?

ANS: 

## Summary of the inspections data

What we have learned about the inspections data? 

* With one or two exceptions, the records are at the business_id and inspection-date level.
* There are inspections for several years.
* Many restaurants have multiple inspections in a year.

We may want to examine the relationship between the scores when a restaurant has multiple inspections in a year.

Next, we will roll some of the information about the inspections up to the business/restaurant level and join the inspection information with the business dataframe. 


# Joining business and inspection data

Now that we have explored the first two data frames, let's merge them into one. 

## Roll up the inspection data

We begin by creating one record per restaurant in the inspection data.
We continue with the data frame `ins2016`, which contains:

+ inspections for 2016 only
+ new variables: `yr` and `mon` for the year and month of the inspection
+ plus all of the original variables

From `ins2016`, create a new data frame called `ins2016grp` which has:

+ One record for each restaurant.
+ `business_id`: the id of the restaurant
+ `date`: the date of the inspection
+ `score`: the inspection score
+ `mon`: the month that the inspection was made
+ `num_ins`: the number of inspections the restaurant had.

If the restaurant had multiple inspections, then keep only the first score and the first month.

Use `group_by()` and `summarize()` to perform this grouping.
```{r}
ins2016grp = 1

```

Use `checkGran()` to confirm the granularity of the `ins2016grp`.
There should be 4543 unique records.

```{r}
checkGran( )
```


## Joining 

Now that the inspections data has one record per restaurant, merge 
this data frame with `busR`. Recall that `busR` consists of restaurants
in popular areas. 

+ Keep only those restaurants that had an inspection in 2016 and are in popular areas. 
+ Keep only the id, name, address, and zip of a restaurant, as well as the inspection date, score, month of inspection, and number of inspections.

Call this new data frame `ins.bus2016`.

```{r}
ins.bus2016 = 1
```


Use `checkGran()` to confirm the granularity of the `ins.bus2016`.
```{r}
checkGran(ins.bus2016["business_id"])
```
You should now have 4362 unique observations.

## Save this data frame to use in HW 7
```{r}
save(ins.bus2016, file = "rest16.rda")
```




