---
title: 'STAT 33A Lab: R Expressions and Vectors'
author: "YOUR NAME (YOUR SID)"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
Section: Write the time of your section  
GSI: Write the name of your GSI

### Instructions
Write R code in the designated code chunks to answer each of the questions. As you work, you may find it helpful to be able to run your code. You can run a single line of code by pressing Ctrl + Enter. You can run an entire code chunk by clicking on the green arrow in the upper right corner of the code chunk. 

Knit the document from time to time to make sure that your code runs without errors from top to bottom in a fresh R environment. 

The code below controls the number of significant digits shown for the return values in your knitted document.

```{r}
options(digits = 3)
```

Begin by running the following code chunk to load the objects that you will be practicing subsetting with. This code loads the data frame called *family* into your workspace.

```{r}
load(url("http://www.stat.berkeley.edu/users/nolan/data/afamily.rda"))
```

Replace "#CODE" in the code blocks with the appropriate R expressions.

### Question 1
Convert the values in `fheight` into cm and assign the new
measurements to `htCM`.

```{r, question1}
htCM = fheight * 2.54
```

*Check your answer*: If you did this right, you should have obtained
numbers that are roughly: 178 163 185 170 155



### Question 2
Create a logical vector called `bigMen` that has `TRUE` for
males who are over weight.

```{r, question2}
bigMen = (fsex == "m") & foverWt
head(bigMen)
```
*Check your answer*: After creating `bigMen`,  
the first few values in the vector should be
TRUE FALSE FALSE FALSE.


### Question 3

The 5% trimmed mean is obtained by taking the mean of the middle 90% of the
values (trimming off 10% of the values). 
REad the documentation for `mean()` and use this function
to find the 5% trimmed mean of `bmi`.
Assign the return value from the call to `mean()` to
`bmi.mean5`
```{r}
bmi.mean5 = mean(fbmi, trim = 0.05)
bmi.mean5
```


### Question 4
Create a vector called `fweight.ran` that is the sum of
`fweight` and normal random values from a normal distribution
with mean 0 and sd 2.

```{r, question4}
set.seed(37) #do not change this code.
fweight.ran = fweight + rnorm(length(fweight),
                              mean = 0, sd = 2)
head(fweight.ran)
```
The first few values of `fweight.ran` should be 
175.2 124.8 186.2 155.4  96.3

### Question 5
Investigate, coercion. 
Run the following code and inspect the results.
Write an explanation as to why the return values make sense.

```{r, question5}
fsex + fbmi
foverWt * fbmi
c(foverWt, fnames)
c(foverWt, fsex)
```
WRITE YOUR EXPLANATION BELOW:

1. Trying add a factor and a numeric. The underlying values in the factor don't correspond to meaningful numbers so the results are `NA`.

2. foverWt is coerced to 0s and 1s in order to perform the multiplication. The result is a vector that contains bmi for those who are over weight and 0 for those who aren't.

3. When concatenating a logical and a character vector, the logical is coerced to character "TRUE" and "FALSE" 

4.When concatentating a logical and a factor, they are both coerced to numeric. The TRUEs become 1 and the FALSEs 0s. The factor values are converted to the numeric value for the level, 1, 2, 3, ... See below for an explicit coercion of the factor `fsex` to a numeric.

```{r}
as.numeric(fsex)
```


## Creating Vectors


### Question 6
Use `seq()` to create a vector of the even numbers between 2 and 20. Assign this to `x`

```{r}
x = seq(from = 2, to = 20, by = 2)
x
```

### Question 7 
Examine the pattern in the values below and create the vector `y` with these values. Do not simply use concatenation.
3 5 9 17 33

```{r}
y = 2^(1:5) + 1
y
```


### Question 8 
Create a logical vector that is TRUE if a weight in `fweight` is within 40 pounds of 150. Call this vector `weight.150` 

```{r}
weight.150 = abs(fweight - 150) <= 40
head(weight.150)
```


Once you have completed these tasks, check that you have written your name at the top of this document and that the document can be knitted.

Finally, it's good practice to print the output of `sessionInfo()` so that others who want to reproduce your results know what packages and version of R you were using when you first did your analysis. 

```{r, sessionInfo}
# leave this as-is:
sessionInfo()
```

Submit the Rmd file to bcourses. Do Not submit the knitted file.

