---
title: 'Lab #2: Three ggplot Plots'
output:
  pdf_document: default
---


```{r, echo=FALSE}
require(dplyr)
require(ggplot2)
```

Run this chunk only once to install LaTeX, which will allow you to knit the document to PDF. 
```{r eval = F}
# remove eval = F from the following line to install
# run this chunk only once
install.packages('tinytex')
tinytex::install_tinytex()
```

## Prepare

We will use the babies data for this labn.

```{r}
load(url("http://www.stat.berkeley.edu/users/nolan/data/babies.Rda"))
```

Your task is to make 3 plots. We have provided written descriptions
of what each plot is expected to display. Be sure to review the lecture handout for tips on how to make plots. Also, you may want to 
search online for how to make some of plots.


Like in class, you will examine just the babies born to mothers who never smoked and mothers who smoked while pregnant. We use the `filter()` function in the `dplyr` package to select the rows.
Change the x-axes scale so that the smallest observation is cut out of
the picture.

```{r}
babies2 = filter(babies, smoke == "Never" | smoke == "Current")
```


## Plot #1

Make a scatter plot of father's height against mother's height.
Jitter the points so that they do not overlap as much.
Add meaningful axis labels and title to the plot.

```{r}

```

## Plot #2

Make a histogram of parity. Use varying bin sizes.
In particular, use bins of width one centered on the 
integers for most of the bins, and then use two wider bins
for the largest values. Change the color of the outline of the
bars so that they are distinct from the fill color of the bars,
and use alpha transparency to lighten the fill color.



```{r}

```

## Plot #3

Make a plot with two overlaid density curves of birthweight, i.e.,
nne density curve for current smokers and one for never smokers.
Use color to differentiate the curves.
Change the title to the legend and the labels for the colors in
the legend to be more informative.


```{r}

```


## Extra Plot for Extra Credit

If you want to try another plot, just for fun (and to earn one point of extra credit), make a two-by-two grid  
of four plots, one for each level of mothers education (drop the mothers with an NA value for education or those with a trade edcuation). 

Make scatter plots of birthweight on mother's height and:

+ Jitter the points. 
+ Use color to distinguish the smokers from never smokers
+ Add regression lines for smokers and never smokers to each plot.
+ Make the points slightly transparent so it's easier to see over plotting
+ Change the x-axes scale so that the smallest observation is cut out of
the picture.

Be sure to read the information about facets at the end of the
class notes.

```{r}

```

## TURN IN A PDF TO GRADESCOPE
