---
title: "Lab 04: Reshaping Data"
subtitle: "STAT 33A"
author: "<<<YOUR NAME HERE>>>"
date: "10/11/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = FALSE}
load(url("http://www.stat.berkeley.edu/users/nolan/data/toyboat.rda"))
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Instructions
    - If you install any packages in the Rmd file, make sure you remove the code to install them before you submit the lab. 
    - Make sure the data is loaded via the URL prior to submitting the lab. 
    - When replacing "<<<YOUR ANSWER HERE>>>" make sure you remove all the angle (<, >) brackets
    - Upload a PDF of your knitted document to gradescope.

# Exercise 1
## Merging Data Sets

Run this chunk below that uses the `rbind()` function. 
```{r}
all_sailors = rbind(sailor1, sailor2)
all_sailors
```


### Are the observations/entries in the dataframe above unique?
<<<YOUR ANSWER HERE>>>


Create a new dataframe called `unique_sailors` that uses `sailor1` and `sailor2`, but doesn't include duplicates.

```{r}
unique_sailors = union(sailor1, sailor2, by = "sid")
unique_sailors
```
There should be 5 vrows in this data frame, one each for dustin, lubber, rusty, yuppy and guppy.

# Exercise 2
## Merging Data Frames

Create `reservations_with_sailors` that appends sailor's names from `unique_sailors` to the `reservations` dataframe. 

```{r}
reservations_with_sailors = 
  right_join(select(unique_sailors, sid, sname), 
             reservations, by = "sid")
reservations_with_sailors
```

To the `reservations_with_sailors` dataframe you created above, append the boat names.

```{r}
boats.sailors = left_join(reservations_with_sailors,
                          select(boat, bid, bname), 
                          by = "bid")
boats.sailors
```
The resulting data frame should have 20 rows. 
Four of these rows appear below:
```{r, eval = FALSE}
# sid   sname   bid day bname
# 31	lubber	101	19	Interlake
# 31	lubber	101	20	Interlake
# 58	rusty	104	20	Marine
# 58	rusty	102	21	Interlake
```


# Exercise 3
## Subsetting

Subset the excursions in `boats.sailors` that used a boat called "Interlake". 

```{r}
interlake = filter(boats.sailors, bname == "Interlake")
interlake
```
The resulting data frame should have 13 rows.


## Exercise 4
Create a data frame that contains the names and ids of the sailors who rented a red boat and the boat names and ids of the boats they rented.  

Call this data frame `red_boat_rental`.
This data frame should have 5 variables: `sid`, `sname`, 
`bid`, `bname`,  and `color`. There should be one record for each sailor, boat combination.

```{r}
red_boat_rental = left_join(
  select(boats.sailors, sid, sname, bid), 
  select(boat, bid, bname, color), by = "bid")

red_boat_rental = filter(red_boat_rental, color == "red")
red_boat_rental = group_by(red_boat_rental, sid, bid)
red_boat_rental = summarize(red_boat_rental, sname = first(sname),
                            bname = first(bname), 
                            color = first(color))

red_boat_rental
```
The resulting data frame should have 3 rows. One record in this data frame is
```{r, eval = FALSE}
# sid bid sname   bname   color 
# 22  104 dustin  Marine  red
```
