---
title: 'STAT 33A Lab 5: Writing Functions'
author: "YOUR NAME (YOUR SID)"
output:
  pdf_document: default
---

## Introduction

In this lab, we will start writing functions and explore:

1. How to return a value from the function call
2. Functions with required, default, and no arguments
3. How to check the inputs to a functionn


## Return value from a function call

We begin with examining 2 very simple functions, `getZero1` and `getZero2`.

```{r}
getZero1 = function(x) {
  return(0)
}

getZero2 = function(x) {
  0
}
```

What is the difference between these 2 functions? Call each of these functions with the following inputs, i.e., `x`, `y`, and `z`. Compare the return values.

```{r}
x = 1:5
y = letters
z = c("this", "is", "a", "test")

getZero1(x)
getZero1(y)
getZero1(z)
getZero2(x)
getZero2(y)
getZero2(z)
```

In `getZero2` there is no explicit return statement so the function returns the result of the last performed evaluation. Therefore the outputs of these two functions are identical.

## A function with no arguments

Q1. Write a new function `getZero3` which takes no arguments and returns the same value as `getZero1`.

```{r}
getZero3 = 1

```

Test your function.

```{r, error = TRUE}
getZero3()
```

## A function with multiple arguments

Q2. Now write a function called `getNumPower` that takes a numeric vector and returns a vector of the same length where each element is a square of the corresponding element of the input vector. Call the input argument `x` and make it a required argument.

```{r}
getNumPower = 1

```

Test your function with the following calls.

```{r, error = TRUE}
getNumPower(1:5)
getNumPower(integer(0))
getNumPower(-20)
```

Note that the input can be a vector of length 1 or 0.

Q3. Modify your function to take a second argument, called `power`, and use this argument to raise the values in `x` to the provided power. Provide a default value to `power` so that it returns the same results as the previous version of the function when the `power` argument is not supplied.

```{r}
getNumPower = 1

```

Note that we can supply a vector for `power` so each element in `x` is raised to a different power. Convince yourself that this is the case by examining the return values from the following calls.

```{r, error = TRUE}
getNumPower(x = x)
getNumPower(1:6, power = rep(0:1, each = 3))
getNumPower(1:6, power = 1:3)
```

Notice that in the call `getNumPower(x = 1:6, power = 1:3)` the input for `x` is a vector of length 6 and the input for `power` is of length 3. What did our function return? Can you reason out why?

Try the following call. Notice it issues a warning. Why? Does it still return a value? Can you figure out what `R` is doing here?

```{r, error = TRUE}
getNumPower(x, power = 1:4)
```

These function calls show how the recycling rule works in `R` when operations are performed on two vectors that have different lengths.

## Practical functions
Write a function called `tip_calc` that will help you calculate the appropriate tip to leave when you eat out at a restaurant. It should take the following arguments: `wo_tax` (pre-tax total), `tip_perc`(percent tip). 
The function can assume that the value for tip_perc is a fraction, e.g.,
0.18 to leave an 18% tip.


```{r}
tip_calc = 1
```

```{r,error=TRUE}
tip_calc(15, 0.15)    # This should output 2.25.
```


Create a new function called `predict_grade` that will help you predict what percent you need on your final exam to have a certain overall grade in the course. Remember [RogerHub](https://rogerhub.com/final-grade-calculator/)? 
$$\frac{d - c(1 - w)}{w}$$

  + c = Current overall grade in class, excluding undetermined categories
  + d = Desired minimum overall grade in class
  + w = Weight of the final exam category



Your function should take three arguments (similar to RogerHub): 

+ current = your current grade (a score between 0 and 100)
+ desired = the grade you want in the course (a score between 0 and 100),
+ weight = the weight of the final exam category (a fraction), e.g., 0.10 if the final is 10% of the grade. 

All of these parameters are required.


```{r}
predict_grade = 1

```

```{r,error=TRUE}
predict_grade(92.4, 90, .10)    # Test this outputs 68.4
predict_grade(88, 90, 0.15)     # Returns 101.33
```
