---
title: 'EDA: Bike Sharing'
author: "lab template"
output:
  pdf_document: default
---

```{r, collapse=TRUE}
require(readr)
require(ggplot2)
require(tidyr)
require(dplyr)
```


## Introduction

*REMEMBER TO KNIT TO PDF AND UPLOAD TO GRADESCOPE*

If you need help knitting to PDF, visit the GSIs office hours and/or attend lab session. 

*KNIT THIS DOCUMENT NOW - AT THE START OF YOUR WORK TO MAKE SURE THAT YOU CAN KNIT TO PDF*

Load the Rda file in to your R session.
It contains two dataframes. The one called `bike' is at the hourly level and the one called `bike_day' is aggregated to the daily level. 
These dataframes contain all of the new and derived variables from the earlier section.

```{r}
load(url("https://www.stat.berkeley.edu/~nolan/Pretoria/data/bikeshare.rda"))
```


### 'bike_day'

The variables in the 'bike_day' data frame are defined as:

* `registered': total number of registered riders for each day
* `casual': total number of casual riders for each day
* `weekday': day of the week (Sun, Mon, ...) 
* `workingday': whether that day is a working day or not (yes, no)
* `holiday': whether that day is a holiday or not (yes, no)
* `season': season (spring, summer, fall, winter)
* `mnth': month
* `temp': average hourly temperature for the day

  
### 'bike'

The granularity of this data frame is at the hourly level.
Each record represents one hour of a day in 2011-2012.

  - instant : integer representing hour and day as an integer
  - season : season (1:spring, 2:summer, 3:fall, 4:winter)
  - yr : year (0: 2011, 1:2012)
  - mnth : month ( 1 to 12)
  - hr : hour (0 to 23)
  - holiday : whether day is holiday or not
  - weekday : day of the week
  - workingday : if day is neither weekend nor holiday
  - weathersit : 
      - 1: Clear or partly cloudy 
      - 2: Mist + clouds
      - 3: Light Snow or Rain
      - 4: Heavy Rain or Snow
  - temp : Normalized temperature in Celsius (divided by 41)
  - atemp: Normalized feeling temperature in Celsius (divided by 50)
  - hum: Normalized percent humidity (divided by 100)
  - windspeed: Normalized wind speed (divided by 67)
  - casual: count of casual users
  - registered: count of registered users
  - cnt: count of total rental bikes including casual and registered
  - nweDate : the data in a special date format
  


## Plot #1

Use `bike_day` to plot two over laid histograms of registered and casual rider counts, where the histograms are distinguished by colore and the bars are filled with somewhat transparent color
so that we can see where the two histograms overlap.  

```{r}


```


## Plot #2

Use the `bike_day` data frame to examine the relationship between casual and 
registered riders.
To address the issue of overplotting, instead of plotting points (`casual` againt `registered`),
use a kernel density plot. Use the  `geom_density_2d()` geom.

```{r}


```



## Plot #3

In class we "stacked" the `bike` data frame:

```{r}
bikeStack =  gather(bike, "rider", "count", casual, registered)
```



Lastly, use `bikeStack` to examine how the rider counts vary by time of day.
Use `geom_smooth()` to smooth the counts over hour of the day.
Produce comparisons of the casual and registered riders.

```{r}


```

