---
title: 'STAT 33A Lab 7: Creating checkGran() function from HW6'
author: "YOUR NAME (YOUR SID)"
output:
   pdf_document: default
---

## Working with `duplicated()`

In this lab, we will work with `duplicated()` and `unique()` and `which()`, which will help us with HW6.

The R function `duplicated()` returns a logical vector where TRUE specifies which elements of a vector or dataframe are duplicates.
```{r,error = TRUE}
duplicated(c(1,2,3,4,1,2,3,4))
```

Next we use duplicated with a dataframe.
Run the following code to create a dataframe called `df`.  
Examine `df` to determine if it has any duplicate rows. 
```{r}
df = data.frame(column1 = c(1, 1, 2,2), column2 = c("x","x","y","z"), column3 = c("a","a","b","b"))
df
```
Here, row 2 is a duplicate of row 1.
The duplicated function checks the entire row for duplicates and returns a logical vector the same length as the number of rows in the dataframe.
```{r}
duplicated(df)
```

When we sum the results from the call to `duplicated()`, we get 1 as there is only 1 duplicate row.
```{r}
sum(duplicated(df))
```
Now that we know that the data has one duplicate row, we might use the `which()` function to figure out which row in the data frame is the duplicate. 

## Working with `which()`

The which function takes the input of logical vector and returns the indices where it's TRUE.

```{r}
which(c(FALSE,FALSE,TRUE,TRUE))
```

Use `which()`, to figure out which of the rows in `df` is the duplicate.
```{r}

```
The answer should be 2 as only the second row is duplicate in our dataframe. 
(We don't count the first row as a duplicate because it is the first occurrence
of the values: 1, "x", "a")

## Creating `checkGran()`

Use all that we have learnt about these functions to write the checkGran() function.

Step1: Create a function named `checkGran()` which has arguments `key` and `dups`, and where `dups` has a default value of FALSE.
```{r}
checkGran=function(){
  
}
```

Step 2: In the function, calculate number of rows in `key` and assign it to a variable `num`. 
Create another variable `dup_location` which returns a boolean vector indicating whether each row in `key` is duplicate or not.
```{r}
checkGran=function(){
  num=1
  dup_location=1
}
```

Step 3: Now, use the logical vector `dup_location` and count the number of unique records and assign it to the new variable `uniq`. Please note that the values in `dup_location` are TRUE if the element is a duplicate record and FALSE if it's not a duplicate.
```{r}
checkGran=function(){
  num=1
  dup_location=1
  uniq=1
}
```

Step 4. Create a vector `info` containing `num` and `uniq`.
```{r}
checkGran=function(){
  #num=1
  #dup_location=1
  #uniq=1
  info=c(num=num, uniq=uniq)
}
```

Write an if statement which updates `info` by including the indices of duplicate elements in the dataframe, if the argument `dups` is TRUE.

```{r}
checkGran=function(){
  num=1 #number of rows in keys
  dup_location=1
  uniq=1 #number of unique rows in keys
  info=c(num=num, uniq=uniq)
  if(dups){
    #add to the info vector the indices of duplicate rows
  }
}
```

Step 5. Lastly, write an explicit return statement at the end of your function to return `info`.
```{r}
checkGran=function(){
  num=1 #number of rows in keys
  dup_location=1
  uniq=1 #number of unique rows in keys
  info=c(num=num, uniq=uniq)
  if(dups){
    #add to the info vector the indices of duplicate rows
  }
  #Your return statement goes here
}
```

Please note - This is one way of writing your `checkGran()` function. There are other approaches to writing it that are also correct.

HW 6 has some additional checks for your function.




  



