---
title: "Getting Started with R"
output:
  pdf_document: default
  word_document: default
---


## Expressions in R

There are 2 basic kinds of expressions:

* Arithmetic

```{r}
2 + 3
```

* Function-style (aka Call Expressions)

```{r}
sample(7)
```

Function-style expressions can have a side effect, 
such as making a plot

```{r}
plot(pressure)
```

R has may built-in functions. Some are arithmetic functions, such as `sqrt`, `log`, and `sin`.  Others perform statistical computations, such as `mean`, `median`, `sd`, and `range`.

## Order of Operation

Arithmetic Expressions follow the typical order of operations:
What is the return value of the following expression?
```{r, eval=FALSE}
3 + 2 * 6 / 3 * (1 + 3)
```

1. 4
2. 10
3. 19
4. 40
5. None of the above

## Read - Evaluate - Print - Loop (REPL)

R is an interactive language. It enables exploratory data analysis. 
We type an expression at the prompt and hit the return / enter key. 
Then R **Reads** this expression, **Evaluates** it (performs a computation), and **Prints** the return value at the console.  
It is then ready for another expression, i.e., another **Loop**.


## Assignment

As just mentioned, when we evaluate an expression, *R* prints the results to the console as output.

```{r}
10 ^ (15 / 3 - 2)
```

If we want to save the return value to use in another calculation, we assign it to a **variable**. For example,

```{r}
x = 10 ^ (15 / 3 - 2)
```

Now, when we run this code, nothing is printed to the console. However, the variable `x` contains the value `1000`.  If we want to confirm this, we can type `x` at the prompt and the value is printed at the console.

```{r}
x
```

We can also use `x` in another computation, e.g.,

```{r}
sqrt(x)
```

Variables have a **name** and a **value**.
To access the value we use the name.
Variables allow us to:

* Store a value without needing to re-compute it
* Write a general expression, e.g., `sqrt(a^2 + b^2)`
* Reduce redundancy (and mistakes)

### Copying

*R* uses "copying" semantics in assignments. 
That is, when we assign the value of `x` to `y`, 
then `y` gets the value of `x`, but the variable
`y` is not linked to `x`. 
This means that when `x` is changed, `y` does not see that change. 
In the code below, `x` begins with the value of 3, then *R* copies this value in the assignment statement so that `y` has the value of 3. These two variables are unrelated after that so when we
assign `x` the value of 10, `y` remains unchanged. Below is the code for this simple example,
```{r}
x
y = x
x
y
```
Again, `y` continues to have the value 3 after `x`'s value has changed.

```{r}
x = 10
x
y
```


Consider the following expressions.
```{r}
z = 2
x = z * y
y = x
```

What are the values in `x`, `y`, and `z` after evaluating the above expressions?
```{r}
x
y
z
```

## Compound Expressions

Compound expressions occur when we compose functions, i.e., we provide the return value from one function as the input for another.

Below are a few examples.
```{r}
abs(z - x)

sqrt(abs(z - x))

round(sqrt(abs(z - x)))

round(sqrt(abs(z - x)), digits = 1)
```


## Help with Function Calls

To learn how to invoke a function, we can read the documentation in R. 
For example, `?head` or `help(head)` opens up help pages for the `head` function in the Help window.

```{r, eval=FALSE}
?head
```

There, we read that `head` has 2 arguments, `x`  and `n`. This function call will result in the help tab displaying information that looks like the following:

```{r,eval=FALSE}
head(x, ...)
## Default S3 method:
head(x, n = 6L, ...)
## S3 method for class 'data.frame'
head(x, n = 6L, ...)
```

The `n` argument has a default value of 6, which means that we don't need to specify the value for this argument, if the default value
is what we want.  No default is provided for `x` so we must provide a value for this argument when we call the function. 

 
Let's try calling `head` with the `precip` object

```{r}
head(precip)
```

The return value is the first 6 values in `precip`. 
We can print the first 3 elements of `precip` instead, by specifying that the argument `n` is 3. We can do this with any of the
following function calls

```{r, collapse=TRUE}
head(precip, 3)
head(precip, n = 3)
```

R is very flexible in allowing us to call the function in many different ways. If we do not provide the argument names then it uses the sequence of the values in the function call to assign these values to the arguments.  If we use an argument name, then the value provided
is assigned to that named argument, e.g., with `head(precip, n = 3)` 
the argument `n` is assigned the value of 3 and the unnamed argument value `precip` is matched by position to the remaining argument(s), i.e., to `x`.  

It's a good idea to use argument names. Typically, we don't provide
the name of the first argument, which is often required, and we do provide the names for the remaining arguments. In other words, the typical function call is `head(precip, n = 3)`.

There are a few more details about calling functions, which we will return to later.




