---
title: "Data Types and Vectors"
output:
  word_document: default
  pdf_document: default
---


## Assignment

Recall, if we want to save the return value to use in another calculation, we assign it to a **variable**. For example,

```{r}
x = 10 ^ (15 / 3 - 2)
```

Now, when we run this code, nothing is printed to the console. However, the variable `x` contains the value `1000`.  If we want to confirm this, we can type `x` at the prompt and the value is printed at the console.

```{r}
x
```

We can also use `x` in another computation, e.g.,

```{r}
sqrt(x)
```

Variables have a **name** and a **value**.
To access the value we use the name.
Variables allow us to:

* Store a value without needing to re-compute it
* Write a general expression, e.g., `sqrt(a^2 + b^2)`
* Reduce redundancy (and mistakes)


## Data Types

R has a number of built-in data types.  The three most basic types are numeric, character, and logical vectors. 
There is no such thing as a scalar in *R*.

A vector is an *ordered* container *primitive elements* of the same type.

We have data on the weather in Port Elizabeth. These are monthly figures for minimum temperature, maximum temperature, precipitation, hours of sunshine (8 or more hours a day), water temperature (comfort lavel), and month. What are the data types?

We load the data with 
```{r}
load("../data/PEweather.rda")
#load(url(
#  "http://www.stat.berkeley.edu/users/nolan/data/PEweather.rda"))
```

The function `ls()` shows us the contents of our work space.
```{r}
ls()
```

Let's take a look at some of these variables:
```{r}
mon
maxTemp
precip
sun8
swim
```
There are some obvious differences between them.
We will explore more about these differences.

We can examine the data types and the summary statistics for each of these by calling `class()` and `summary()`, e.g.,
```{r}
class(precip)
summary(precip)
```

What do you think a summary of `mon` will give?
```{r}
summary(mon)
```


Let's look at `class(sun8)`
```{r}
class(sun8)
summary(sun8)
```
Notice that the `summary()` function provides different responses depending on the type of data. In the case of logical vectors, it gives a simple count of the number of TRUEs and FALSEs.

A factor vector is a special storage class used for qualitative data.
The values are internally stored as integers, but each integer corresponds to a level or categoy. 
The categories have labels that are character strings.
```{r}
class(swim)
levels(swim)
summary(swim)
```

How many different versions of the return value of the `summary()` function have you seen? 

We have examined logical, character, numeric, and factor types. Do they each have a different summary?

## Return Values from Function Calls 

In *R* we have 

+ aggregator functions such as `min()`, `mean()`.
+ functions that operate on vectors and return vectors of the same length, such as `abs()`
+ functions that return different types of objects, e.g. `summary()` and `lm()` 

Here are examples of each. 

### Aggregator Functions

We first call a few aggregator functions:
```{r}
min(maxTemp)

mean(maxTemp)

sd(maxTemp)
```

### Functions that Return Vectors

Here are some examples of functions that return vectors the same length as the input vector.

Convert `precip` from milimeters to centimeters:

```{r}
precip / 10
```

Sort `maxTemp`: 
```{r}
sort(maxTemp)
```

Determine which entries in `maxTemp` are below 21:
```{r}
maxTemp < 21
```

### Functions that Return Special Objects

Examples of functions that return different sorts of objects follow:

The `summary()` function returns a table with 6 entries
```{r}
y = summary(precip)
class(y)
y
```

The `plot()` function has a plot as a side effect
```{r}
plot(maxTemp, type ="l")
```

The `lm()` function fits a least squares line to data and return an `lm` object,
which is a kind of list that has a lot of information relevant to the fit.
```{r}
lm.out = lm(precip ~ maxTemp)
class(lm.out)
names(lm.out)
```
Don't worry if you don't know what a least squares line is.
We're only using it as an example.

We can call `plot()` with the `lm` object.
We can call `summary()` with the `lm` object.
We can call `predict()` with the `lm` object.
And, we can access various elements of the `lm` object to further
analyze the result of the linear fit.

```{r, FALSE}
plot(lm.out)
summary(lm.out)
```

## Logical Operations

Often we want to compare the elements in a vector to see if they are equal to, greater than, less than, etc. to a particular value.

For example, we can determine how many months have at least 250 hours of sunshine
```{r}
over250 = sunshine >= 250
over250
```

We can find how many months have warm water for swimming 
```{r}
swim == "warm"
```

Two aggregator functions that are useful with logical vectors are `all()` and `any()`. The first returns `TRUE` if all elements in
the logical vector or `TRUE`. The `any()` function returns `TRUE` if and of the elements in the input vector are `TRUE`.
```{r}
all(over250)
any(over250)
```
We see that not all of the months have at least 250 hours of sunshine, but some of them do.

## Boolean Operations

At times we want to operate on two or more logical vectors.
For example, we may want to determine whether ther are months with
over 8 hours of sunshine and warm water. 
```{r}
sun8 & (swim == "warm")
```

Or, we may want to know which months have over 8 hours of sunshine or warm water:
```{r}
sun8 | (swim == "warm")
```


## Creating Vectors

It can be useful to create vectors to assist with statistical analyses. We demo some of the ways to do this.

The simplest approach is to create vectors whose elements are 1-apart.
```{r}
1:5
-10:-7
```

The `seq` function is more sophisticated version of `:`
```{r}
seq(from = 0, to = 100, by = 10)
seq(from = 0, to = 100, length = 5)
seq(from = 0, by = 10, length = 4)
```

There is also the `c()` function, which we can use to concatenate values together into a vector:
```{r}
x = c(1, 20, 0, NA, Inf)
x
```

We can concatenate vectors together into one long vector, e.g.,
```{r}
c(minTemp, maxTemp)
```

And the `rep()` function. Can you figure out what each is doing? 
```{r}
rep(x, each = 2)
rep(x, times = 2)
```
And
```{r, eval=FALSE}
rep(x, times = c(3, 1, 5, 0, 2))
```


## Special Values

We put a couple of special characters into the `x` vector that we created earlier. 
Let's see what happens when we work with them. 

First, let's add 2 to each element of `x`. Before we do this, what do you think might happen when we try to add 2 to an `NA` value?  `NA` stands for Not Available. If we add 2 to a value
that is not available, then is the new value still not available?
What about adding 2 to infinity?
```{r}
2 + x
```
Does the return value make sense?

What do you think the average of the elements in `x` should be?
9? or NA? or Inf? or an error?
```{r}
mean(x)
```

What happens when we compare `x` to 37, i.e., which elements in `x` are less than 37.
```{r}
x < 37
```


## Vectorized Computations

As we described early, *R* works with vectors.
We can add and subtract vectors and multiply and divide them.

We provide some examples.
We can find the differernce between the max and min monthly temperatures with 
```{r}
maxTemp - minTemp
```

We can also multiply minTemp and precip (which makes no sense)
```{r}
minTemp * precip
```


## Coercion 

What happens in the following multipliclation
```{r}
precip * sun8
```
Here We have asked *R* to multiply a numeric vector with a logical vector.
*R* coerces the logical vector into a numeric vector of 0s and 1s to perform the operation.

Cercing a logical vector to 0s and 1s can be very useful.
For example, describe in words what each of the return values is
for the following function calls
```{r}
sum(sun8)

sum(maxTemp * sun8) / sum(sun8)
```


What happens in the following concatentation:
```{r}
c(maxTemp, mon)
```
This time *R* coerced the ages into character strings.

A vector must have homogenous elements.
How does *R* decide which data type to coerce?
Well, *R* can't coerce "Jan" or "Mar" into a number so it
chooses to coerce 25 and 24 into strings.

## Recycling

What is happening in the following:
```{r}
precip
precip + (1:6)
```
*R* *recycles* the values in the shorter vector to match the length of the longer vector and then performs the summation.


What happens when the shorter vector is not an integer multiple of the longer one?
Try it and find out

```{r}
precip + (1:7)
```

It performs the computation for us, but gives a warning message.






