---
title: 'Intro to ggplot'
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

## Prepare

If you haven't done so already, install the ggplot package (and dependencies). Then, load it into your workspace with:
```{r}
require(dplyr)
require(ggplot2)
```

We will use the babies data for this demonstration.

```{r}
load("../data/babies.Rda")
```

We examine just the babies born to mothers who never smoked and mothers who smoked while pregnant. We use the `filter()` function in the 
`dplyr` package to select the rows.

```{r}
babies2 = filter(babies, smoke == "Never" | smoke == "Current")
```


## The ggplot model


### The ggplot object

A call to `ggplot()` creates a plot object, which it need not print. Below we assign a call to `ggplot()` to the variable `bwtPlot`.

```{r}
bwtPlot = ggplot()
```

Notice we have no plot as a side effect.
We created the ggplot object, but we didn't even give it any data.

### The Data

ggplot only accepts data in data frames. If you have two vectors that you want to make a scatter plot from, then you must put them into a data frame.

### The Aesthetic Mapping

The aesthetic mapping connects variables in a data frame to a feature of the plot, such as `x` and `y` locations, `color`, and `size`. As an example, the aesthetic: `aes(x = ht, y = bwt, color = smoke)` for the babies data, maps mother’s height to the x-axis, birth weight to the y-axis, and uses color to denote smoking status.

### The Layer

A plot is made in layers. Each layer must have a data frame and aesthetic mapping (with a few exceptions).

We can build the plot by adding layers to the plot object.

For example, we can make a scatter plot by adding a layer of points to the plot. We do this with:
```{r fig.width=5,fig.height=3}
bwtPlot + geom_point(data = babies2, 
                 mapping = aes(x = ht, y = bwt, color = smoke))
```

It's that easy!
Notice the `x` axis is mother's height, the `y` axis is 
birthweight, and the points are colored according to the mother's smoking status. The aesthetic mapping was used to specify 
the points, i.e., the mapping gave the x-coordinate, y-coordinate, and color 
of each poing

### More Layers

We can add more layers to the plot. That's the paradigm.
For example, we can add fitted regression lines, 
one for each group of mothers (current and never smokers) with:

```{r fig.width=5,fig.height=3}
bwtPlot + 
  geom_jitter(data = babies2, 
             mapping = aes(x = ht, y = bwt, color = smoke)) +
  geom_smooth(data = babies2, 
              mapping = aes(x = ht, y = bwt, color = smoke, 
                            group = smoke),
              method='lm', formula=y~x, se = FALSE)
```


### Specifying Data Frame and Aesthetic Mapping in the plot object

Our code can be greatly simplified, if we specify the data frame
and the aesthetic mapping in the original call to `ggplot()`.
Then, we don't need to specify it in each of the layers.
Even when specified in a `ggplot()`, it is still possible to 
override this specification in a layer.

```{r, fig.width=5,fig.height=3}
ggplot(data = babies2, 
       mapping = aes(x = ht, y = bwt, color = smoke)) +
  geom_point() +
  geom_smooth(method='lm', se = FALSE)
```

Notice that in the `geom_smooth()` layer we used the 
mapping specified earlier in `ggplot()` and we added one more
mapping of the group aesthetic to smoking status. 
This aesthetic indicates that we want the smooth to be done in groups
so that we have separate lines for the never and current smokers.

## Some Plot Examples

There are many kinds of plots that we can make with `ggplot()`.
We show a few of the basics here:



### Histogram

```{r, fig.width=5,fig.height=3}
ggplot(data = babies2, 
       mapping = aes(x = bwt)) +
  geom_histogram(aes(y = ..density..)) 
```


### Box plot

```{r, fig.width=5,fig.height=3}
ggplot(data = filter(babies2, !is.na(ed)), 
       mapping = aes(x = ed, y = bwt)) +
  geom_boxplot() 
```


### Bar Chart

```{r, fig.width=5,fig.height=3}
ggplot(data = babies2, 
       mapping = aes(x = ed)) +
  geom_bar(fill="palegreen", color = "blue")
```

One habit that ggplot has that I find annoying is that it creates bars and boxes for NA values. It's a bit trickly to eliminate them, but possible. There are a few ways to do it. One is to filter the data frame.

```{r, fig.width=5,fig.height=3}
ggplot(data = filter(babies2, !is.na(ed)), 
       mapping = aes(x = ed)) +
  geom_bar() 
```

## Beyond the Basics 

### Changing Defaults

There are many aspects to the plot that we can change.
These include:

+ colors for non-data mappings
+ plotting symbol
+ plotting size 

Here's one plot that changes many of the defaults: 

```{r, fig.width=5,fig.height=3}
ggplot(data = babies, aes(x = bwt)) +
  geom_histogram(aes( y = ..density..), 
                 breaks=c(50, seq(70, 170, by=2), 180) ,
                 alpha = 0.4, col = "red")
```

*IMPORTANT*: the color argument in the `geom_histogram()` function call is not the same as the color argument to the aesthetic call `aes()`. The color specification of red simply says that we want
to change the appearance of the histogram. It is not doing any aesthetic mapping.


### Scales

The `scale_XXX()` functions control the scales of data components in a plot. That is, they connect the scale of the data values to
the aesthetic space. 

These scale layers include:

+ x and y axis scale for specifyig the limits and labels for axes
+ color palettes for the color scale

These specifications are observed in the plot via the axes and legends.
 
Below we adjust all three scales: the `x`, `y`, and `color`.
For the `x` scale, we provide a title (name) for the axis and the 
endpoint limits.
For the color scale, we provide a different palette (black and red) and more informative labels for these colors.

```{r, fig.width=5,fig.height=3}
ggplot(data = babies2, aes(x = ht, y = bwt, color = smoke)) +
  geom_jitter(width = 0.5, alpha = 0.8, na.rm = TRUE) +
  scale_x_continuous(name = "Mother's Height (in)", limits = c(58, 71)) +
  scale_y_continuous(name = "Birth Weight (oz)") +
  scale_color_manual(values = c("black", "red"),
                     labels = c("Never Smoker", "Current Smoker"))
```

Changing the axis labels and plot title are so common that short cut functions that are simpler to call have been provided.
See for example, the plot below where we specify the axis labels and plot title with a call to `labs()` rather than `scale`.

```{r, fig.width=5,fig.height=3}
ggplot(data = babies2, aes(x = ht, y = bwt, color = smoke)) +
  geom_jitter(width = 0.5, size = 0.75, alpha = 0.8, na.rm = TRUE) +
  geom_smooth(se = FALSE) +
  scale_x_continuous(limits = c(58, 71)) +
  scale_color_manual(values = c("black", "red"),
                     labels = c("Never Smoker", "Current Smoker")) +
  labs(x="Mother's Height (in)", y="Birth Weight (oz)",
       title="Scatter Plot", color="Smoking\nStatus")
```



### Themes

The `theme()` changes the over appearance of the plot.
That is, it customizes the non-data components of the plot,
such as titles, labels, fonts, background gridlines, legends,
and more. 

Themes can give plots a consistent customized look to your plots.
It is possible to change small aspects of the appearance of a plot, or to adopt a "whole new" theme. 

Below we change the 
theme to one that is good for black and white renderings.
We save the current (active) theme as we set the new one, and then
we restore the old one after making the plot.

```{r, fig.width=5,fig.height=3}
ggplot(data = babies, aes(x = bwt)) +
  geom_histogram(aes( y = ..density..), 
                 breaks=c(50, seq(70, 170, by=2), 180) ,
                 alpha = 0.4, col = "red") +
  theme_bw()
```
Notice how the gray background disappeared, and the grid lines are 
pale grey rather than white.

The `theme()` function gives us a lot of control in tweaking the
appearance of the non-data components. For example, we can change 
the font family, font color, and type of the legend text.
Below is a plot that uses `theme()` to control the placement of the 
legend. It also adopts the "classic" theme so that the plot looks like a base R plot.

```{r, fig.width=5,fig.height=3}
ggplot(data = babies2, aes(x = ht, y = bwt, color = smoke)) +
  geom_jitter(width = 0.5, size = 0.75, alpha = 0.8, na.rm = TRUE) +
  theme(legend.position = c(0.1, 0.9)) +
  theme_classic()
```

Changing the axis labels and plot title are so common that short cut functions that are simpler to call have been provided.
See for example, the plot below where we augment the above plot with
labels.

```{r, fig.width=5,fig.height=3}
ggplot(data = babies2, aes(x = ht, y = bwt, color = smoke)) +
  geom_jitter(width = 0.5, size = 0.75, alpha = 0.8, na.rm = TRUE) +
  theme(legend.position = c(0.1, 0.9)) 
```

### Annotations

The `annotate()` layer does not behave like the geom layers. Specifically, `annotate()` does not map variables in a data frame to aesthetics. Instead, the input arguments are provided via vectors. This kind of layer is useful for adding fixed reference information to plots.

Below we add a vertical line to the plot (note that it too doesn' use the data frame) and we annotate this line with text.

```{r, fig.width=5,fig.height=3}
ggplot(data = babies, aes(x = bwt)) +
  geom_histogram(aes( y = ..density..), alpha = 0.4, col = "black") + 
  labs(x = "Birth Weight (ounces)") +
  geom_vline(xintercept = 88, lwd = 2.5, col = "red") +
  annotate(geom = "text", x = 80, y = 0.02, 
           label = "Low\n Birth\n Weight", size = 3)
```




### Facets

We can use `facet()` to juxtapose plots of subsets of the data.
That is, we want panels of the same type of plot for different 
groups of data, where these groupos are determined by categories of a variable.
This sort of plot is called a facet plot. 

Below we make two panels, one for each type of smoker.
Each panel is a scatter plot of birth weight on height.
Notice the code is very similar to an earlier scatter plot
that superposes these two groups of mothers and uses color to
distringuish between them.

```{r, fig.width=6,fig.height=3}
bwtPlot = ggplot(data = babies2, 
             mapping = aes(x = ht, y = bwt)) + 
            geom_jitter(width = 0.5, size = 0.75, 
                        alpha = 0.8, na.rm = TRUE) + 
            geom_smooth(method='lm', formula=y~x, se = FALSE)

bwtPlot + facet_wrap(~ smoke)
```

