---
title: 'EDA: Bike Sharing'
author: "An Example"
output:
  html_document: default
  word_document: default
---

```{r, collapse=TRUE}
require(readr)
require(ggplot2)
require(tidyr)
require(dplyr)
```

## Introduction

In this notebook, we work with a dataframe called `bike', which is at the hourly level. We load it into our *R* session with

```{r}
load(url("https://www.stat.berkeley.edu/~nolan/Pretoria/data/bikeshare.rda"))
```

  
### 'bike'

Each record represents one hour of a day in 2011-2012.

  - instant : integer representing hour and day as an integer
  - season : season (1:spring, 2:summer, 3:fall, 4:winter)
  - yr : year (0: 2011, 1:2012)
  - mnth : month ( 1 to 12)
  - hr : hour (0 to 23)
  - holiday : whether day is holiday or not
  - weekday : day of the week
  - workingday : if day is neither weekend nor holiday
  - weathersit : 
      - 1: Clear or partly cloudy 
      - 2: Mist + clouds
      - 3: Light Snow or Rain
      - 4: Heavy Rain or Snow
  - temp : Normalized temperature in Celsius (divided by 41)
  - atemp: Normalized feeling temperature in Celsius (divided by 50)
  - hum: Normalized percent humidity (divided by 100)
  - windspeed: Normalized wind speed (divided by 67)
  - casual: count of casual users
  - registered: count of registered users
  - cnt: count of total rental bikes including casual and registered
  - nweDate : the data in a special date format
  


## Exploring the Distribution of Riders

Begin by comparing the distributions of the hourly counts of casual and registered riders. Make density plots of these two types of riders.

Note that we need to overlay the histograms of the variables `casual` and `registered`. We do this by adding two geom_density layers, one
for casual and one for registered riders.

```{r}
ggplot(data = bike) +
  geom_density(aes(x = casual, y = ..density..), bw = 5, col = "red") + 
  geom_density(aes(x = registered, y = ..density..),
              bw = 5, col = "blue") +
  labs(x = "Hourly Count of Riders")
```
*The distribution of hourly counts of casual riders has a sharp peak near 0, and is skew right with a long right tail. The distribution of registered riders also has a peak at 0, though much less sharp. The distribution has a large shoulder from about 50 to 250, which could be a second mode.*


The density plots do not show us how the counts for registered and casual riders vary together. To see this relationship, we can make a scatter plot. We map the variable `workingday` to the color aesthetic to distinguish the points according to whether the day is a weekend/holiday or not. 

```{r}
ggplot(data = bike, 
       aes(x = casual, y = registered, color = workingday)) +
  geom_point(size = 0.4, alpha = 0.4) +
  labs(x = "Casual Riders (hourly counts)",
       y = "Registered Riders (hourly counts)")
```

*This scatter plot of counts of the 2 types of riders has points color coded according to whether or not the day is a weekday. We see that there are two distinct elliptical shapes, one for week days and one for week ends. On the weekends, the casual and registered riders have a roughly linear relationships. On the other hand, during the work week, there appears to be no relationship between these counts. The numbers of casual riders in an hour fall between 0 and 100, and for any given count of casual riders, the registered riders ranges from about 0 to 750, regardless.* 

## Gathering counts

It will be more convenient to put the causal and registered rider counts together into one variable and create a new variable that denotes which type of rider the count corresponds to.  The `gather()` function in the `dplyr` package is handy for doing this.
We stack the counts in `registered` and `casual` into a new variable called `count`  and create a new variable for the type of rider called `rider`. 

```{r}
bikeStack =  gather(bike, "rider", "count", casual, registered)
```


## Exploring Ride Sharing and Time

We narrow down the range of time to the month of June in 2011, and make a line plot using color to distinguish between the 2 types of riders.

```{r rideTime}
bikeJune = filter(bikeStack, mnth == 6 & yr == 0)

ggplot(data = bikeJune) +
  geom_line(aes(x = newDate, y = count, color = rider)) +
  labs(x = "Date", y = "Number of Riders")
```

This plot has several interesting features. How do the number of casual and registered riders compare for different times of day and days of the week? 

*We can see several things in this line plot. We see daily patterns in rentals, with the weekends having far fewer registered users and more casual riders. The number of casual riders is always less than the number of registerred riders, but on the weekend, the counts are much closer together. We can also see 3 peaks in the registered riders daily weekday schedule. These look to be in the morning, mid-day, and evening.*

Next, we make boxplots to examine the relationship between time and rider counts.

```{r myRideTimePlot1}
ggplot(data = bikeStack) +
  geom_boxplot(aes(x = factor(mnth), y = 1 + count, fill = rider)) +
  scale_y_log10() + labs(x = "Month", y = "Number of Riders")
```

*These monthly side-by-side boxplots for casual and registered riders shows a greater difference in counts of casual riders in the spring-fall months in comparison to the winter months than for the registered riders.*

Next we examine the ridership by the time of day. We plot only smooths of the counts.

```{r myRideTimePlot2}
ggplot(data = bikeStack, aes(x = hr, y = count, color = rider)) +
  geom_smooth() +
  labs(x = "Hour of Day", y = "Number of Riders")
```

*These smooth curves for rider counts for casual and registered riders have several interesting features. The 3 peaks that we saw earlier in the line plot for registered riders appear here as a peak at morning and evening commute hours and another smaller peak at lunch time.  On the other hand, the casual riders have a higher usage after 10 a.m. with a broad flat peak between noon and 5 p.m. Notice that the curve for registered users dips below 0 at 4 a.m. This is due to the bandwidth of the curve being fitted to the data - the curve is a locally piecewise cubic and in order for it to rise rapidly for the morning rush hour, it dips below 0 in the early morning hours.*

We can also look at scatter plots, rather than smooth curves.

```{r myRideTimePlot3}
ggplot(data = bikeStack, aes(x = hr, y = count, color = rider)) +
  geom_point(size = 0.4, alpha = 0.2, position = "jitter") +
  labs(x = "Hour of the Day", y = "Number of Riders (in an hour)")
```

*This scatter plot is a graph of the same data as in the previous plot. The difference is that we are plotting the individual points rather than showing the locally smoothed curve. It's evident here that for each hour, the casual riders has a long-tailed distribution of counts, but given the scale of the peaks for the registered riders, it's hard to see what's happening for the casual riders.*

## Exploring Ride Sharing and Day of week

Next we examine how the weather is associated with rider's behavior. First, we look at how the proportion of casual riders changes with the temperature.

We create a new variable called *propCasual* that is the proportion of casual riders among total riders in an hour. Add this variable to the data frame *bike*.  Make a scatter plot of *propCasual* and *temp*. The temperature should be jittered to avoid over plotting. Also, we color the points according to the day of the week.

```{r exploreWeather}
bike = mutate(bike, propCasual = casual/(casual + registered))

ggplot(data = bike) +
  geom_point(aes(x = temp, y = propCasual, color = weekday), 
             position = "jitter", size = 0.4, alpha = 0.5) +
  labs(x = "Temperature (normalized to 0-1)",
       y = "Proportion of Casual Riders")
  
```

*We see the proportion of casual riders is higher on the weekends. That there are many days where this proportion is basically 0, for all temperatures. We also see that for temperatures above 0.75, the proportion of casual riders appears larger.*

Many points are overlapping in this scatter plot, and although the plot shows trends, they are difficult to verify. As an alternative, we can fit smooth curves to show the average proportion of casual ridership in time, separately for each day of the week. 

We again use the `geo_smooth` layer to average the `propCasual` values for each temperature value. In the call to `geom_smooth()` we use the `formula` argument to specify that we want the smooth curve to be a cubic spline.
 
```{r}
ggplot(data = bike) +
  geom_smooth(aes(x = temp, y = propCasual, color = weekday),
              method = lm, formula = y ~ splines::bs(x, 3), se = FALSE) +
  labs(x = "Temperature (normalized to 0-1)",
       y = "Proportion of Casual Riders")
```

*As temperature increases, there is a roughly linear increase in the proportion of casual riders. The wekeends have higher proportions of casual riders in the warmer temperatures.*

## Exploring Ride Sharing and Weather

The following set of boxplots examines the proportion of casual riders by weather.  

```{r myRideWeatherPlot1}
ggplot(data = bike) +
  geom_boxplot(aes(x = weathersit, y = propCasual)) +
  labs(x = "Weather Situation",
       y = "Proportion of Casual Riders")
```

*In these side-by-side box plots of the proportion of riders for different types of weather, we see that for the days when the weather is poor, the distribution of casual riders is shifted downward (more for rain than for mist).*

We see that there are few days with heavy wether so we probably want to drop it from this plot.

```{r myRideWeatherPlot2}
ggplot(filter(bike, weathersit != "Heavy"), 
       aes(x = newDate, y = propCasual, color = weathersit)) +
  geom_smooth(se = FALSE)  +
  labs(x = "Date", y = "Proportion of Casual Riders")
```

*These 3 smooth curves show the changes in proportion of casual riders for clear, misty, and light precipitation days throughout the year. These curves rise and fall together with clear and misty days being close to each other with misty typically below clear, and the light rain/snow days are consistently below the other 2 types of days.  We also see the proportion of casual riders increase in the summer months. This proportion appears larger in 2011 than 2012. This may be due to the program gaining ridership among the registered users from 2011 to 2012.*

Lastly, we examine the relationship between the proportion of casual riders and the temperature. To do this, we bin temperature and create a boxplot for each bin.  Note that we use the `group` aesthetic for the binning, rather than the `fill` aesthetic. If you swap group to fill, you will want to suppress the legend.


```{r}
ggplot(data = bike) +
  geom_boxplot(aes(x = temp, y = propCasual,
                   group = cut_width(temp, 0.05))) +
  labs(x = "Temperature (normalized to 0-1)",
       y = "Proportion of Casual Riders")
```

*This last set of boxplots show a flattening of the proportion in warm weather.*
