---
title: "Simulations and the Urn Model"
output:
  pdf_document: default
  word_document: default
---

## Roulette 

The game of roulette involves spinning a wheel that has 38 slots, numbered 1 through 36 and 0 and 00. Also the 0 and 00 slots are green and the remaining 36 slots are evenly divided between red and black. 
A ball is tossed/released onto the spinning wheel, it bounces around, and eventually settles into one of the 38 slots.  

There are many types of bets you can place in the game. 
In this exercise, we compare two types of bets and we also look at what happens when we bet different amounts.  

1. Bet on Red: Betting on the ball landing in a red slot pays even odds. This means that if you bet $1 on Red and the ball lands on Red then you keep your $1 and the casino pays out $1. If the ball lands in any other color slot then you lose your $1.  

2. Bet on #17: Betting on a particular number pays 35 to 1 so if you place a $1 bet on 17 and the ball lands on it, you keep your dollar and the casino pays out $35. If ball lands in any other slot then you lose your $1 bet.

Of course you need not bet only $1, you may bet a larger amount. We will examine what happens when you place many $1 bets or fewer $100 bets.

### Urn Model for a bet on red

Create an urn that describes the possible outcome (gain or loss) from a $1 be on red.

* How many marbles are in the urn?
* What are the different values that appear on the marbles?
* How many marbles have each value?

Construct a vector, called `urn` that matches the answer to these questions.

```{r}
urn = c(rep(-1, 20), rep(1, 18))
```

Use the `sample()` function to simulate 2,000 bets of $1 on red.
Be sure to specify the arguments:

+ x
+ size
+ replace

Assign the return value from `sample()` to `samps`.

```{r}
samps = sample(urn, size = 10000, replace = TRUE)
```


On average, what is the net gain of one bet?
What is the sd of one bet?

```{r}
mean(samps)
sd(samps)
```

The expected gain is negative. We can expect to lose about 5 cents on every dollar we bet.  This is because there are 18 ways for the ball to land Red and 20 for it to land Black or Green, but the Bet is paid like it's fair (1 to 1), meanng the casino has an advantage.

The SD is very close to 1. This makes sense because the two outcomes are +1 and -1 and the mean is close to 0 so the deviations are all a little more than 1 or a little less than 1.

### Urn Model for a bet on red

Create a new urn called `urn17` that represents the win and loss from a $1 bet on number 17.  Simulate this bet 2000 times (assign the results to `samps17`) and find the average and SD of the results.

  
```{r}
urn17 = c(rep(-1, 37), rep(35, 1))

samps17 = sample(urn17, size = 10000, replace = TRUE)

mean(samps17)
sd(samps17)
```

Again the expected value is small but negative. In fact, we expect to lose about a nickel on every $1 bet.
However, the SD is about 5.7 for this bet.

## Repeated bets 

Suppose you have $100, and you are considering whether to place:

  * 100 $1 bets on Red
  * 100 $1 bets on 17
  * 1 $100 bet on Red
  * 1 $100 bet on 17
  
For each of these options simulate 10,000 rounds and assign the outcome to samps100.red, samps100.17, samps1big.red and samps1big.17.

###  Simulate 100 $1 bets on Red

We will get you started. For each round of 100 bets, we are interested in the total winnings, i.e., the sum of the net gain.
Below we compute this quantity for one round

```{r}
sum(sample(urn, 100, replace = TRUE))
```

Now use the `replcate()` function to repeat this 10,000 times.
```{r}
samps100.red = replicate(10000, 
                         sum(sample(urn, 100, replace = TRUE)))
```


###  Simulate 100 $1 bets on #17

```{r}
samps100.17 = replicate(10000, 
                         sum(sample(urn17, 100, replace = TRUE)))
```

###  Simulate 1 $100 bet on red

Note that we simply need to scale the urn we created and make one bet.

```{r}
urn100.red = 100 * urn
samps1big.red = replicate(10000, 
                  sum(sample(urn100.red, 1, replace = TRUE)))
```

### Simulate 1 $100 bet on #17


```{r}
urn100.17 = 100 * urn17
samps1big.17 = replicate(10000, 
                  sum(sample(urn100.17, 1, replace = TRUE)))
```


### Compare the average gain

```{r}
mean(samps100.red)
mean(samps100.17)
mean(samps1big.red)
mean(samps1big.17)
```

All four types of bets have the same expected value: -$5.26.
Your simulation might show slight deviations from this because afterall it is a simulation.

### Compare the standard deviation in gain

```{r}
sd(samps100.red)
sd(samps100.17)
sd(samps1big.red)
sd(samps1big.17)
```

These betting schemes have very different SDs. The smallest SD is when we bet $1 at a time on Red. The fluctuations around -$5 are about $10.  

The other SDs are much bigger. Can you reason why?

Placing a single $100 has a much larger SD than 100 smaller bets. If we place it on Red then the SD is about $100. This makes sense because we are either going to win $100 or lose $100 in one bet, and the chances of winning/losing are nearly equal.

The single $100 bet on number 17 has the largest SD: over $570. We have a small chance of getting $3500 and a large chance of losing $100. Those are the only possible outcomes. 

### Compare their probability histograms 

```{r}
bets = data.frame(gains = c(samps100.red, samps100.17, 
                            samps1big.red, samps1big.17),
                  type = rep(c("red.1", "No17.1", "red.big", "no17.big"), each = length(samps100.red)))
```


```{r}
plot(table(samps100.red), 
     xlab = "Net Gain", ylab = "Count", 
     main ="Place $1 on Number 17 for 100 Bets",
     type = "h")

```


What do you see in these plots that you don't see by simply computing the expected value and SD of this random process?  

The probability histogram for the net gain from the 100 bets of $1 on Red looks approximately normal.  The histogram for the 100 bets on 17 looks skew right, here the chance of a win is small and there are a large number of bets.  
Finally, the histograms for the single $100 bets show that there are  only two possible outcomes.

### Different goals 

Use  the simulation results to estimate the chance that you:

  * Lose less than $10
  * Win more than $400
  
```{r}
n = length(samps100.red)
sum(samps100.red > -10 )/n
sum(samps100.17 > -10)/n
sum(samps1big.red > -10 )/n
sum(samps1big.17 > -10)/n

sum(samps100.red > 400 )/n
sum(samps100.17 > 400)/n
sum(samps1big.red > 400 )/n
sum(samps1big.17 > 400)/n
```

For the two bets on Red and the 100 bets on #17, the chance of losing less than $10 ranges between 47% and 65%.  The $100 bet on 17 only has a chance of about 2.5% of losing less than $10.

On the other hand, in the 10,000 simulations of these three bets (the two Red bets and the 100 bets on #17), none of them ended up with a net gain greater than $400.  
In fact, it's not possible in the two types of bets on Red to win more than $100.  
Howver, if you win at all in the $100 bet on #17, then you win big so your chance is also 2.5% that you win more than $400.
