---
title: "33A Intro to ggplot"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
require(dplyr)
```

## Prepare

We will use the babies data for this demonstration.

```{r}
load(url("http://www.stat.berkeley.edu/users/nolan/data/babies.Rda"))
```

We examine just the babies born to mothers who never smoked and mothers who smoked while pregnant. We use the `filter()` function in the 
`dplyr` package to select the rows.

```{r}
babies2 = filter(babies, smoke == "Never" | smoke == "Current")
```


## The ggplot model

If you haven't done so already, install the ggplot package (and dependencies). Then, load it into your workspace with:
```{r}
require(ggplot2)
```

### The ggplot object

A call to `ggplot()` creates a plot object, which it need not print. Below we assign a call to `ggplot()` to the variable `bwtPlot`.

```{r}
bwtPlot = ggplot()
```

Notice we have no plot as a side effect.
We created the ggplot object, but we didn't even give it any data.

### The Data

ggplot only accepts data in data frames. If you have two vectors that you want to make a scatter plot from, then you must put them into a data frame.

### The Aesthetic Mapping

The aesthetic mapping connects variables in a data frame to a feature of the plot, such as `x` and `y` locations, `color`, and `size`. As an example, the aesthetic: `aes(x = ht, y = bwt, color = smoke)` for the babies data, maps mother’s height to the x-axis, birth weight to the y-axis, and uses color to denote smoking status.

### The Layer

A plot is made in layers. Each layer must have a data frame and aesthetic mapping (with a few exceptions).

We can build the plot by adding layers to the plot object.

For example, we can make a scatter plot by adding a layer of points to the plot. We do this with:
```{r}
bwtPlot + geom_point(data = babies2, 
                 mapping = aes(x = ht, y = bwt, color = smoke))
```

It's that easy!
Notice the `x` axis is mother's height, the `y` axis is 
birthweight, and the points are colored according to the mother's smoking status. The aesthetic mapping was used to specify 
the points, i.e., the mapping gave the x-coordinate, y-coordinate, and color 
of each poing

### More Layers

We can add more layers to the plot. That's the paradigm.
For example, we can add fitted regression lines, 
one for each group of mothers (current and never smokers) with:

```{r}
bwtPlot + 
  geom_point(data = babies2, 
             mapping = aes(x = ht, y = bwt, color = smoke)) +
  geom_smooth(data = babies2, 
              mapping = aes(x = ht, y = bwt, color = smoke, 
                            group = smoke),
              method='lm', formula=y~x, se = FALSE)
```


### Specifying Data Frame and Aesthetic Mapping in the plot object

Our code can be greatly simplified, if we specify the data frame
and the aesthetic mapping in the original call to `ggplot()`.
Then, we don't need to specify it in each of the layers.
Even when specified in a `ggplot()`, it is still possible to 
override this specification in a layer.

```{r}
ggplot(data = babies2, 
       mapping = aes(x = ht, y = bwt, color = smoke)) +
  geom_point() +
  geom_smooth(mapping = aes(group = smoke),
              method='lm', formula=y~x, se = FALSE)
```

Notice that in the `geom_smooth()` layer we used the 
mapping specified earlier in `ggplot()` and we added one more
mapping of the group aesthetic to smoking status. 
This aesthetic indicates that we want the smooth to be done in groups
so that we have separate lines for the never and current smokers.

## Some Plot Examples

There are many kinds of plots that we can make with `ggplot()`.
We show a few of the basics here:

### Histogram

```{r}
ggplot(data = babies2, 
       mapping = aes(x = bwt)) +
  geom_histogram() 
```


### Box plot

```{r}
ggplot(data = babies2, 
       mapping = aes(x = ed, y = bwt)) +
  geom_boxplot() 
```


### Bar Chart

```{r}
ggplot(data = babies2, 
       mapping = aes(x = ed)) +
  geom_bar() 
```

One habit that ggplot has that I find annoying is that it creates bars and boxes for NA values. It's a bit trickly to eliminate them, but possible. There are a few ways to do it. One is to filter the data frame.

```{r}
ggplot(data = filter(babies2, !is.na(ed)), 
       mapping = aes(x = ed)) +
  geom_bar() 
```


