---
title: "Reshaping Data"
author: "lab template"
output:
  pdf_document: default
  word_document: default
---

```{r, collapse=TRUE}
require(readr)
require(ggplot2)
require(tidyr)
require(dplyr)
```


## Introduction

Load the Rda file in to your R session.
It contains the dataframe called `bike' which holds counts of riders at the hourly level.

```{r}
load(url("https://www.stat.berkeley.edu/~nolan/Pretoria/data/bikeshare.rda"))
```

  
### 'bike'

Each record represents one hour of a day in 2011-2012.

  - instant : integer representing hour and day as an integer
  - season : season (1:spring, 2:summer, 3:fall, 4:winter)
  - yr : year (0: 2011, 1:2012)
  - mnth : month ( 1 to 12)
  - hr : hour (0 to 23)
  - holiday : whether day is holiday or not
  - weekday : day of the week
  - workingday : if day is neither weekend nor holiday
  - weathersit : 
      - 1: Clear or partly cloudy 
      - 2: Mist + clouds
      - 3: Light Snow or Rain
      - 4: Heavy Rain or Snow
  - temp : Normalized temperature in Celsius (divided by 41)
  - atemp: Normalized feeling temperature in Celsius (divided by 50)
  - hum: Normalized percent humidity (divided by 100)
  - windspeed: Normalized wind speed (divided by 67)
  - casual: count of casual users
  - registered: count of registered users
  - cnt: count of total rental bikes including casual and registered
  - newDate : the data in a special date format
  


## Filter Rows and Select Columns

Let's work with a subset of the data:

+ records from one year (2012)
+ columns registered, casual, working day, week day, hour, month, and the new date

Call this new data frame bike12.

First let's reduce the columns, and reassign to itself.
We use `select()` to reduce the columns. 

```{r}
keepVars = c("yr", "registered", "casual", "workingday", 
             "weekday", "hr", "mnth", "newDate")
bike = select(bike, keepVars)

names(bike)
```

Note that we kept `yr` because we will need it for the next step.

Before we take the subset of rows, let's find the counts so we can do a before-after comparison to check our work.

```{r}
table(bike$yr)
```

Now let's take the subset.

```{r}
bike12 = filter(bike, yr == 1)

nrow(bike12)
```

We have the correct number of rows. We do not have the correct columns.
We do not want `yr`. We can use `select()` to drop it.


```{r}
bike12 = select(bike12, -"yr")

names(bike12)
```

This is an unusual way to deselect a column; we used a negative sign on 
a character vector.

Let's make a plot to celebrate:

```{r}
ggplot(data = bike12, aes(x = hr, y = casual)) +
  geom_smooth() +
  geom_smooth(aes(y = registered), color = "green") +
  labs(x = "Hour of Day", y = "Number of Riders")
```


## Gather the registered and casual counts

Recall, that we would rather make these plots with a stacked dtata frame.
This is because `registered` and `casual` both have the same information in them.
They both contain hourly counts. The difference is that one contains 
counts for registered riders and the other holds counts for casual riders. 

We can use `gather()` to create a new vector called `counts` that contains 
hourly counts for both kinds of riders. At the same time, we need to create
a new variable, `rider` which contains the information as to the type of rider the
count pertains. The `rider` vector is a factor that has two levels: 
casual and registered. 


We make this data frame as follows

```{r}
bikeStack =  gather(bike12, "rider", "counts", casual, registered)
```

Let's examine `bikeStack`. 

+ It should have twice the number of records as `bike12`. 
+ There should be an equal number of registered and casual values in `rider`.
+ A variable, such as `hr`, should have twice the number of records for each hour as in `bike12.

We check these with:

```{r}
2 * nrow(bike12)  == nrow(bikeStack)

table(bikeStack$rider)

all(table(bike12$hr) * 2 == table(bikeStack$hr))
```


Let's remake the plot above, using this new data frame.
`geom_smooth()` to smooth the counts over hour of the day.
Produce comparisons of the casual and registered riders.

```{r}
ggplot(data = bikeStack, aes(x = hr, y = counts, color = rider)) +
  geom_smooth() +
  labs(x = "Hour of Day", y = "Number of Riders")
```

## Group the hourly data to daily data

Suppose we want to examine the data after it has been aggregated to the 
daily counts. We work with `bikeStack`.

What do we group by?

+ hour
+ date
+ rider type
+ hour and date
+ hour and rider type
+ rider type and date

### Obtaining the date

Before we group the data, we have an issue that needs addressing.
The dates include the hour of the day. We just want the date so that 
we can use it as a grouping variable.
Working with dates can be tricky. We use the `lubridate` package,
which has a function `yday()` which returns an integer for the calendar 
day of the year.
```{r}
require(lubridate)
bikeStack = mutate(bikeStack, day.yr = yday(newDate))
```

Let's see how well we did:
```{r}
head(table(bikeStack$day.yr), 12)
```

Does this make sense? Should we have 48 records? Why?

Let's table the table to see for how many days do we have 48 records.

```{r}
table(table(bikeStack$day.yr))
```

What shall we do with these data?
Let's drop the days with fewer than 44 records. Just not yet!

### Grouping by date and rider

```{r}
bs.gdr = group_by(bikeStack, day.yr, rider)
```


Next we create the variables of interest from the groups:

+ sum the `counts` for each day
+ keep  `weekday`, `workingday`, and `mnth` (these should have the same values within a group)
+ drop `newDate` and `hr`


```{r}
bike_day = summarise(bs.gdr, daily = sum(counts), 
                     records = n(), 
                     mnth = first(mnth), 
                     weekday = first(weekday), 
                     workingday = first(workingday))
```

If we want to keep these other variables we must include them in the `summarize()` call. We are simply taking the first value in each group. The values in a group
should all be the same.

Now we also have the number of records in each group so that we can filter out the weird months.

```{r}
bike_day = filter(bike_day, records > 20)
```


Let's make a plot to celebrate!

```{r}
ggplot(data = bike_day, 
       aes(x = factor(mnth), y = daily, color = rider)) +
  geom_boxplot() +
  labs(x = "Month", y = "Daily Ridership", title = "2012",
       color = "Rider\nType")
```


```{r}
ggplot(data = bike_day, 
       aes(x = factor(weekday), y = daily, color = rider)) +
  geom_boxplot() +
  labs(x = "Day of Week", y = "Daily Ridership", title = "2012",
       color = "Rider\nType")
```

