---
title: "Toy Boat Exercises"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
load(url("http://www.stat.berkeley.edu/users/nolan/data/toyboat.rda"))
library(dplyr)
library(tidyr)
library(ggplot2)
```

There are four data frames: sailor1, sailor2, boat, reservations.
We show each as they are all small/toy examples
```{r}
sailor1
```

```{r}
sailor2
```

```{r}
boat
```

```{r}
reservations
```

### Ex 1. Subset columns

Create a subset of `sailor2` that contains only the `age` and `sid` columns.
```{r}
sub1 = select(sailor2, age, sid)
sub1
```


### Ex 2. Subset rows

Create a subset of `sailor2` that contains the names of sailors who
have a rating higher than 8.
```{r}
sub2 = select(filter(sailor2, rating > 8), sname)
sub2
```

## Joining and Grouping Data Frames

For each of the following exercises, list what data frames need to be combined and sketch what the resulting data frame should look like,
in particular what are the number of rows in the resulting data 
frame and the main variables.

Break the task into sub-tasks and use the verbs, select, filter, union, intersect, left_join, inner_join, group_by, and summarize to describe each action. Also specify the variables that will be used in these actions.



### Ex 3. Union
Combine the two sailor data frames (i.e., `sailor1` and `sailor2`) 
into one data frame. Each sailor should appear only once.
```{r}
sailor3 = union(sailor1, sailor2, by = "sid")
sailor3
```

## Ex 4. Joining data frames 
Create a data frame that consists of the names of the sailors that are in `sailor1` and not in `sailor2`.
```{r}
sailor1not2 = setdiff(sailor1, sailor2, "sid")
sailor1not2
```


## Ex 5. Joining and subsetting
Create a data frame of the reservations of boat 103 that includes the sailor's name.
```{r}
reserve103 = filter(reservations, bid == 103)
reserve103 = right_join(select(sailor3, sid, sname), 
                        reserve103, by = "sid")

reserve103
```



## Ex 6. Joining data frames 
Create a data frame of the boats rented by sailors with high ratings (over 8). There should be one record per boat.
```{r}
bestRenters = filter(sailor3, rating > 8)
bestRenters = left_join(bestRenters, reservations, by = "sid")
bestRenters = group_by(bestRenters, bid)
bestRenters = summarize(bestRenters, n = n())
bestRenters
```

## Ex 7. Aggregating rows
Create a data frame of the boat id, name, and the counts of the number of times each boat has been rented (no need to worry about boats that were not rented).
```{r}
rentalCount = left_join(reservations, boat, by = "bid")
rentalCount = group_by(rentalCount, bid)
rentalCount = summarize(rentalCount, bname = first(bname), n = n())
rentalCount
```


## Ex 8. Joining three data frames
Create a data frame that contains the sailor name, boat name, and the number of times the sailor has rented the boat.
```{r}
favoriteRentals = left_join(reservations, 
                            select(boat, "bid", "bname"), 
                            by = "bid")
favoriteRentals2 = left_join(favoriteRentals, 
                            select(sailor3, "sid", "sname"), 
                            by = "sid")

favoriteRentals3 = group_by(favoriteRentals2, bid, sid)

favoriteRentals = summarize(favoriteRentals3, n = n(), 
                            boat = first(bname),
                            sailor = last(sname))

favoriteRentals
```
