---
title: "Dirty Dozen"
output:
  word_document: default
  html_document: default
---

## Dirty Dozen Code Recommendations

The following is set up
```{r}
n = 2000
x = vector(mode="numeric", length = n)
y = x
bigDF = data.frame(x = rnorm(n), y = rnorm(n), z = runif(n))
w = runif(n, min = -10, max = 4)
v = vector(mode = "numeric", length = n)
K = 22
```


1. Name objects with similar, misleading, or uninformative names, such as `x`, `tmp`.

```{r}
z1 = rnorm(10)
z2 = rnorm(10)
z3 = rnorm(10)
```


What is a better approach to naming/creating these three objects? Is there a data structure more suitable for assignment of the random values? What if the three calls to `rnorm()` generated a different number of random values? 

```{r}

```


2. Ignore the type of a vector. 

```{r}
a =  letters[1:8]
a[2] = TRUE
```

Explain why the above code might introduce unexpected behavior later on in your workflow.



3. Make multiple copies of a large object.

```{r}
big1 = bigDF[ , 1]
big2 = bigDF[ , 2]
big3 = bigDF[ , 3]
newDF = data.frame(big1,  diff = big2 - big3)
```

Rewrite the code so that there are no multiple copies of the data.
```{r}

```


4. Initialize a vector, and do it one element at a time. 

```{r}
for (i in 1:n) {
   x[i] = 2
}
```

Create the vector x without a loop (or apply function).
```{r}

```


5. Subset illogically. 

```{r}
for(i in 1:n) {
   if (w[i] < -1) {
     v[i] = 0
   } else {
     if (w[i] > 1) {
       v[i] = 1
     } else v[i] = 3*w[i]^2
   }
}
```


Rewrite this code, eliminating need for loops and if statements.
```{r}

```


6. Rely on globally set variables inside a function. 

```{r}
myFunc = function(x) {
  x = rnorm(100)
  return(x + K)
}
```

Change the signature of the function so that it is clear which external variables are needed.

```{r}

```


7. Apply a vectorized function to a vector one element at a time. 
```{r}
for (i in 1:n){
  y[i] = floor(x[i]) + 2
}
```

Rewrite this code without using a loop or apply function.
```{r}

```


8. Put an operation that needs to be done once inside a loop. 
```{r}
for (i in 2:length(y)){
  k = 2
  y[i] = floor(sqrt(y[i-1])) + k
}
```

Rewrite the code to make it more efficient.
```{r}

```


9. Call a function several times over, each time tweaking the input arguments slightly. 
```{r}
plot(x = seq(0, 0.5, by = 0.02), y = seq(-1.5, -0.5, by = 0.04))
text(0.289, -1.225, "+500,000", cex = 0.3)
text(0.289, -1.245, "+100,000", cex = 0.3)
text(0.289, -1.260, "+50,000", cex = 0.3)
text(0.289, -1.270, "+10,000", cex = 0.3)
text(0.289, -1.280, "+1,000", cex = 0.3)
```

Write a little function, include arguments to text() that do not change from one call to the next. 
Then write an expression to call the function. 
Recall text() is vectorized.
```{r}

```


10.  Keep every possible intermediate result, just in case.
```{r}
z1 = myFunc(x)
z2 = z1 * 2
z3 = log(z2)
z4 = z3 + 10
```


Simplify these computations into one expression.
```{r}

```


11. Leave objects in your workspace rather than saving them to an `rda` file. Save `bigDF` and `myFunc()`to an rda file called `silly.rda` 

```{r}

```


12. Document your code incompletely. 
Keep similar looking functions around with nearly the same names, 
some of which work and some of which don't.
Add documentaion to the following function used in class.

```{r}
funApply = function(x) {
 drop = apply(is.na(x), 1, any)
 x[!drop, ]
}
```

